### 2017-02-13 agutmpl.tex
Initial ShareLaTeX Import

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% AGUtmpl.tex: this template file is for articles formatted with LaTeX2e,
% Modified March 2013
%
% This template includes commands and instructions
% given in the order necessary to produce a final output that will
% satisfy AGU requirements.
%
% PLEASE DO NOT USE YOUR OWN MACROS
% DO NOT USE \newcommand, \renewcommand, or \def.
%
% FOR FIGURES, DO NOT USE \psfrag or \subfigure.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% All questions should be e-mailed to latex@agu.org.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Step 1: Set the \documentclass
%
% There are two options for article format: two column (default)
% and draft.
%
% PLEASE USE THE DRAFT OPTION TO SUBMIT YOUR PAPERS.
% The draft option produces double spaced output.
%
% Choose the journal abbreviation for the journal you are
% submitting to:

% jgrga JOURNAL OF GEOPHYSICAL RESEARCH
% gbc   GLOBAL BIOCHEMICAL CYCLES
% grl   GEOPHYSICAL RESEARCH LETTERS
% pal   PALEOCEANOGRAPHY
% ras   RADIO SCIENCE
% rog   REVIEWS OF GEOPHYSICS
% tec   TECTONICS
% wrr   WATER RESOURCES RESEARCH
% gc    GEOCHEMISTRY, GEOPHYSICS, GEOSYSTEMS
% sw    SPACE WEATHER
% ms    JAMES
% ef    EARTH'S FUTURE
%
%
%
% (If you are submitting to a journal other than jgrga,
% substitute the initials of the journal for "jgrga" below.)

\documentclass[draft,sw]{AGUTeX}

\usepackage{amsmath}
\usepackage{txfonts}
\usepackage{url}
\usepackage{lineno}
 \linenumbers*[1]

%  To add line numbers to lines with equations:
%  \begin{linenomath*}
%  \begin{equation}
%  \end{equation}
%  \end{linenomath*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figures and Tables
%
%
% DO NOT USE \psfrag or \subfigure commands.
%
%  Figures and tables should be placed AT THE END OF THE ARTICLE,
%  after the references.
%
%  Uncomment the following command to include .eps files
%  (comment out this line for draft format):
%  \usepackage[dvips]{graphicx}
%
%  Uncomment the following command to allow illustrations to print
%   when using Draft:
%  \setkeys{Gin}{draft=false}
%
% Substitute one of the following for [dvips] above
% if you are using a different driver program and want to
% proof your illustrations on your machine:
%
% [xdvi], [dvipdf], [dvipsone], [dviwindo], [emtex], [dviwin],
% [pctexps],  [pctexwin],  [pctexhp],  [pctex32], [truetex], [tcidvi],
% [oztex], [textures]
%
% See how to enter figures and tables at the end of the article, after
% references.
%
%% ------------------------------------------------------------------------ %%
%
%  ENTER PREAMBLE
%
%% ------------------------------------------------------------------------ %%

% Author names in capital letters:
\authorrunninghead{CHANDORKAR ET AL.}

% Shorter version of title entered in capital letters:
\titlerunninghead{GAUSSIAN PROCESS $Dst$ MODELS}

%Corresponding author mailing address and e-mail address:
%\authoraddr{Corresponding author: A. B. Smith,
%Department of Hydrology and Water Resources, University of
%Arizona, Harshbarger Building 11, Tucson, AZ 85721, USA.
%(a.b.smith@hwr.arizona.edu)}

\begin{document}

%% ------------------------------------------------------------------------ %%
%
%  TITLE
%
%% ------------------------------------------------------------------------ %%

\title{Gaussian Processes Autoregressive Models for Forecasting the Disturbance Storm Time Index}

\authors{M. Chandorkar,\altaffilmark{1}
 E. Camporeale,\altaffilmark{1}
 S. Wing\altaffilmark{2}}

\altaffiltext{1}{Multiscale Dynamics, Centrum Wiskunde Informatica (CWI), Amsterdam,
              1098XG Amsterdam}

\altaffiltext{2}{The Johns Hopkins University Applied Physics Laboratory, 
              Laurel, Maryland, 20723, USA}

% >> Do NOT include any \begin...\end commands within
% >> the body of the abstract.

\begin{abstract}
(Type abstract here)
\end{abstract}

%% ------------------------------------------------------------------------ %%
%
%  BEGIN ARTICLE
%
%% ------------------------------------------------------------------------ %%

% The body of the article must start with a \begin{article} command
%
% \end{article} must follow the references section, before the figures
%  and tables.

\begin{article}

%% ------------------------------------------------------------------------ %%
%
%  TEXT
%
%% ------------------------------------------------------------------------ %%

\section{Introduction}
(Article text here)

\section{Methodology: Gaussian Process} \label{sec:method}

\emph{Gaussian Processes} first appeared in machine learning research in \citet{Neal:1996:BLN:525544}, as the limiting case of Bayesian inference performed on neural networks with infinitely many neurons in the hidden layers. Although their inception in the machine learning community is recent, their origins can be traced back to the geo-statistics research community where they are known as \emph{Kriging} methods (\citet{krige1951statistical}). In pure mathematics area \emph{Gaussian Processes} have been studied extensively and their existence was first proven by Kolmogorov's extension theorem (\citet{tao2011introduction}). The reader is referred to \citet{Rasmussen:2005:GPM:1162254} for an in depth treatment of Gaussian Processes in machine learning.

Let us assume that we want to model a process in which a scalar quantity $y$ is described by $y = f(\mathbf{x}) + \epsilon$ where   $f(.): \mathbb{R}^d \rightarrow \mathbb{R}$ is an unknown scalar function of a multidimensional input vector $\mathbf{x} \in \mathbb{R}^d$, and $\epsilon \sim \mathcal{N}(0, \sigma^2)$ is Gaussian distributed noise with variance $\sigma^2$.

A set of labeled data points ${(\mathbf{x}_i, y_i); i = 1 \cdots N}$ can be conveniently expressed by a $N \times d$ data matrix $\mathbf{X}$ and a $N \times 1$ response vector $\mathbf{y}$, as shown in equations (\ref{eq:feat}) and (\ref{eq:labels}).

\begin{align}
  \mathbf{X} & = & \left( \begin{array}{c} \mathbf{x}^{T}_1 \\ \mathbf{x}^{T}_2 \\ \vdots \\ \mathbf{x}^{T}_n \end{array} \right)_{n \times d} \label{eq:feat} \\
  \vspace{2\baselineskip}
  \mathbf{y} & = & \left( \begin{array}{c} y_1 \\ y_2 \\ \vdots \\ y_N \end{array} \right) _{n \times 1} \label{eq:labels}
\end{align}

Our task is to infer the values of the unknown function $f(.)$ based on the inputs $\mathbf{X}$ and the noisy observations $\mathbf{y}$. We now assume that the joint distribution of $f(\mathbf{x}_i), i = 1 \cdots N$ is a multivariate Gaussian as shown in equations (\ref{eq:fvalues}), (\ref{eq:normal}) and (\ref{eq:sto}).

\begin{align}
 \mathbf{f} & = & \left( \begin{array}{c} f(\mathbf{x}_1) \\ f(\mathbf{x}_2) \\ \vdots \\ f(\mathbf{x}_N) \end{array} \right) \label{eq:fvalues}\\
 \vspace{2\baselineskip}
 \mathbf{f} | \mathbf{x}_1, \cdots, \mathbf{x}_N & \sim & \mathcal{N}\left( \mathbf{\mu}, \mathbf{\Lambda} \right)  \label{eq:normal}\\
 \vspace{2\baselineskip}
 p( \mathbf{f} \ | \ \mathbf{x}_1, \cdots, \mathbf{x}_N) & = & \frac{1}{(2\pi)^{n/2} det(\mathbf{\Lambda})^{1/2}} exp \left(-\frac{1}{2} (\mathbf{f} - \mathbf{\mu})^T \mathbf{\Lambda}^{-1} (\mathbf{f} - \mathbf{\mu}) \right) \label{eq:sto}
\end{align}

Here $\mathbf{f}$ is a $N\times 1$ vector consisting of the values $f(\mathbf{x}_i), i = 1 \cdots N$. In equation (\ref{eq:normal}), $\mathbf{f}|\mathbf{x}_1, \cdots, \mathbf{x}_N$ denotes the conditional distribution of $\mathbf{f}$ with respect to the input data (i.e., $\mathbf{X}$) and $\mathcal{N}\left( \mathbf{\mu}, \mathbf{\Lambda} \right)$ represents a multivariate Gaussian distribution with mean vector $\mathbf{\mu}$ and covariance matrix $\mathbf{\Lambda}$. The probability density function of this distribution $p( \mathbf{f} \ | \ \mathbf{x}_1, \cdots, \mathbf{x}_N)$ is therefore given by equation (\ref{eq:sto}).

From equation (\ref{eq:sto}), one can observe that in order to uniquely define the distribution of the process, it is required to specify $\mathbf{\mu}$ and $\mathbf{\Lambda}$. For this probability density to be valid, there are further requirements imposed on $\mathbf{\Lambda}$:

\begin{enumerate}
      \item Symmetry: $\mathbf{\Lambda}_{ij} = \mathbf{\Lambda}_{ji} \ \forall i,j \in {1, \cdots, N} $ 
      \item Positive Semi-definiteness: $\mathbf{z}^T \mathbf{\Lambda} \mathbf{z} \geq 0 \ \forall \mathbf{z} \in \mathbb{R}^N$  
\end{enumerate}

Inspecting the individual elements of $\mathbf{\mu}$ and $\mathbf{\Lambda}$, we realise that they take the following form.

\begin{align}
      \mu_i & = & \mathbb{E}[f(\mathbf{x}_i)] := m(\mathbf{x}_i) \\
      \Lambda_{ij} & = & \mathbb{E}[(f(\mathbf{x}_i) - \mu_i)(f(\mathbf{x}_j) - \mu_j)] := K(\mathbf{x}_i, \mathbf{x}_j)
\end{align}

Here $\mathbb{E}$ denotes the expectation (average). The elements of $\mathbf{\mu}$ and $\mathbf{\Lambda}$ are expressed as functions $m(\mathbf{x}_i)$ and $K(\mathbf{x}_i, \mathbf{x}_j)$ of the inputs $\mathbf{x}_i,\ \mathbf{x}_j$. Specifying the functions $m(\mathbf{x})$ and $K(\mathbf{x}, \mathbf{x}')$ completely specifies each element of $\mathbf{\mu}$ and $\mathbf{\Lambda}$ and subsequently the finite dimensional distribution of $\mathbf{f} | \mathbf{x}_1, \cdots, \mathbf{x}_N $. In most practical applications of \emph{Gaussian Processes} the mean function is often defined as $m(\mathbf{x}) = 0$, which is not unreasonable if the data is standardized to have zero mean. \emph{Gaussian Processes} are represented in machine learning literature using the following notation:

\begin{equation}
    f(\mathbf{x}) \sim \mathcal{GP}(m(\mathbf{x}), K(\mathbf{x}, \mathbf{x}'))
\end{equation}

\subsection{Inference and Predictions} \label{sec:inference}

Our aim is to infer the function $f(\mathbf{x})$ from the noisy training data and generate predictions $f(\mathbf{x}^{*}_i)$ for a set of test points $ {\mathbf{x}^{*}_i : \forall i \in 1, \cdots, M} $. We define $\mathbf{X}^*$ as the test data matrix whose rows are formed by $\mathbf{x}^{*}_i$ as shown in equation (\ref{eq:testfeat}). 
\begin{equation}
    \mathbf{X}_* = \left( \begin{array}{c} (\mathbf{x}^{*}_1)^T \\ (\mathbf{x}^{*}_2)^T \\ \vdots \\ (\mathbf{x}^{*}_M)^T \end{array} \right)_{M \times d} \label{eq:testfeat} 
\end{equation}
Using the multivariate Gaussian distribution in equation (\ref{eq:sto}) we can construct the joint distribution of $f(\mathbf{x})$ over the training and test points. The vector of training and test outputs $\left( \begin{array}{c} \mathbf{y} \\ \mathbf{f_*} \end{array} \right)$ is of dimension $(N+M) \times 1$ and is constructed by appending the test set predictions $\mathbf{f}_*$ to the observed noisy measurements $\mathbf{y}$.

\begin{align}
    \mathbf{f}_* & = & \left( \begin{array}{c} f(\mathbf{x^{*}_1}) \\ f(\mathbf{x^{*}_2}) \\ \vdots \\ f(\mathbf{x^{*}_M}) \end{array} \right)_{M \times 1} \\
     \vspace{4\baselineskip}
    \left( \begin{array}{c} \mathbf{y} \\ \mathbf{f_*} \end{array} \right) | \ \ \mathbf{X}, \mathbf{X}_* & \sim & 
    \mathcal{N}\left(\mathbf{0}, \left[ \begin{array}{cc} \mathbf{K} + \sigma^{2} \mathbf{I} & \mathbf{K}_{*} \\ \mathbf{K}_{*}^T & \mathbf{K}_{**} \end{array} \right ] \right) \label{eq:dist}
\end{align}

Since we have noisy measurements of $f$ over the training data, we add the noise variance $\sigma^2$ to the variance of $f$ as shown in (\ref{eq:dist}). The block matrix components of the $(N+M) \times (N+M)$ covariance matrix have the following structure.

\begin{enumerate}
      \item $\mathbf{I}$: The $n \times n$ identity matrix.
      \item $\mathbf{K} = [K(\mathbf{x}_i, \mathbf{x}_j)], \ i,j \in 1,\cdots,n$ : Kernel matrix constructed from all couples obtained from the training data.
      \item $\mathbf{K}_{*} = [K(\mathbf{x}_i, \mathbf{x}^{*}_j)], \ i \in 1,\cdots,n ; j \in 1,\cdots,m$ : Cross kernel matrix constructed from all couples between training and test data points.
      \item $\mathbf{K}_{**} = [K(\mathbf{x}^{*}_i, \mathbf{x}^{*}_j)], \ i,j \in 1,\cdots,m$: Kernel matrix constructed from all couples obtained from the test data.
\end{enumerate}

With the multivariate normal distribution defined in equation (\ref{eq:dist}), probabilistic predictions $f_*$ can be generated by constructing the conditional distribution $\mathbf{f_*}|\mathbf{X},\mathbf{y},\mathbf{X_*}$. Since the original distribution of $\left( \begin{array}{c} \mathbf{y} \\ \mathbf{f_*} \end{array} \right) | \ \ \mathbf{X}, \mathbf{X}_*$ is a multivariate Gaussian, conditioning on a subset of elements $\mathbf{y}$ yields another Gaussian distribution whose mean and covariance can be calculated exactly, as in equation (\ref{eq:posterior}) (see \citet{Rasmussen:2005:GPM:1162254}).
\begin{equation}
    \mathbf{f_*}|\mathbf{X},\mathbf{y},\mathbf{X_*} \sim \mathcal{N}(\mathbf{\bar{f}_*}, \Sigma_*)  \label{eq:posterior},
\end{equation}
where
\begin{align}
    \mathbf{\bar{f}_*} & = & \mathbf{K}^T_{*} [\mathbf{K} + \sigma^{2} \mathbf{I}]^{-1} \mathbf{y} \label{eq:posteriormean} \\
    \Sigma_* & = & \mathbf{K}_{**} - \mathbf{K}^T_{*} \left(\mathbf{K} + \sigma^{2} \mathbf{I}\right)^{-1} \mathbf{K}_{*} \label{eq:posteriorcov}
\end{align}

The practical implementation of \emph{Gaussian Process} models requires the inversion of the training data kernel matrix $[\mathbf{K} + \sigma^{2} \mathbf{I}]^{-1}$ to calculate the parameters of the predictive distribution $\mathbf{f_*}|\mathbf{X},\mathbf{y},\mathbf{X_*}$. The computational complexity of this inference is dominated by the linear problem in Eq. (\ref{eq:posteriormean}), which can be solved via Cholesky decomposition, with a time complexity of $O(N^3)$, where $N$ is the number of data points.

The distribution of $\mathbf{f_*}| \mathbf{X},\mathbf{y},\mathbf{X_*}$ is known in Bayesian analysis as the \emph{Posterior Predictive Distribution}. This illustrates a key difference between \emph{Gaussian Processes} and other regression models such as \emph{Neural Networks}, \emph{Linear Models} and \emph{Support Vector Machines}: a \emph{Gaussian Process} model does not generate point predictions for new data but outputs a predictive distribution for the quantity sought, thus allowing to construct error bars on the predictions. This property of Bayesian models such as \emph{Gaussian Processes} makes them very appealing for Space Weather forecasting applications.

The central design issue in applying \emph{Gaussian Process} models is the choice of the function $K(\mathbf{x}, \mathbf{x}')$. The same constraints that apply to $\mathbf{\Lambda}$ also apply to the function $K$. In machine learning, these symmetric positive definite functions of two variables are known as \emph{kernels}. Kernel based methods are applied extensively in data analysis i.e. regression, clustering, classification, density estimation (see \citet{Scholkopf:2001:LKS:559923}, \citet{hofmann2008}).

\subsection{Kernel Functions}

For the success of a \emph{Gaussian Process} model an appropriate choice of kernel function is paramount. The symmetry and positive semi-definiteness of \emph{Gaussian Process} kernels implies that they represent inner-products between some basis function representation of the data. The interested reader is suggested to refer to \citet{Berlinet2004}, \citet{Scholkopf:2001:LKS:559923} and \citet{hofmann2008} for a thorough treatment of kernel functions and the rich theory behind them. Some common kernel functions used in machine learning include the radial basis function (RBF) kernel $K(\mathbf{x}, \mathbf{y}) = \frac{1}{2} exp(-||\mathbf{x} - \mathbf{y}||^2/l^2)$ and the polynomial kernel $K(\mathbf{x}, \mathbf{y}) = (\mathbf{x}^\intercal \mathbf{y} + b)^d$.

The quantities $l$ in the RBF, and $b$ and $d$ in the polynomial kernel are known as \emph{hyper-parameters}. Hyper-parameters give flexibility to particular kernel structure, for example $d = 1, 2, 3, \cdots$ in the polynomial kernel represents linear, quadratic, cubic and higher order polynomials respectively. The method of assigning values to the \emph{hyper-parameters} is crucial in the model building process.

In this study, we construct Gaussian Process regression models with linear kernels as shown in equation (\ref{eq:usedKernel}), for \emph{one step ahead} prediction of the $Dst$ index. Our choice of kernel leaves us with only one adjustable hyper-parameter, the model noise $\sigma$.

\begin{equation}\label{eq:usedKernel}
    K(\mathbf{x}, \mathbf{y}) = \mathbf{x}^\intercal \mathbf{y}
\end{equation}

We initialize a grid of values for the model noise $\sigma$ and use the error on a predefined validation data set to choose the best performing value of $\sigma$. While constructing the grid of possible values for $\sigma$ it must be ensured that $\sigma > 0$ so that the kernel matrix constructed on the training data is non-singular.

\section{One Step Ahead Prediction} \label{sec:osa}

Below in equations (\ref{eq:Dst}) - (\ref{eq:GPPoly}) we outline a \emph{Gaussian Process} formulation for \emph{OSA} prediction of $Dst$. A vector of features $\mathbf{x}_{t-1}$ is used as input to an unknown function $f(\mathbf{x}_{t-1})$.

The features $\mathbf{x}_{t-1}$ can be any collection of quantities in the hourly resolution OMNI data set. Generally $\mathbf{x}_{t-1}$ are time histories of $Dst$ and other important variables such as plasma pressure $p(t)$, solar wind speed $V(t)$, $z$ component of the interplanetary magnetic field $B_z(t)$.

\begin{align}
    Dst(t) & = & f(\mathbf{x}_{t-1}) + \epsilon \label{eq:Dst} \\
    \epsilon & \sim & \mathcal{N}(0, \sigma^2) \label{eq:GPNoise} \\
    f(x_t) & \sim & \mathcal{GP}(m(\mathbf{x}_t), K(\mathbf{x}_t, \mathbf{x}_s)) \label{eq:DstGP} \\
    K(\mathbf{x}, \mathbf{y}) & = & \mathbf{x}^\intercal \mathbf{y} + b \label{eq:GPPoly}
\end{align}

In the following we consider two choices for the input features $\mathbf{x}_{t-1}$ leading to two variants of \emph{Gaussian Process} regression for $Dst$ time series prediction.

\subsection{Gaussian Process Auto-Regressive (GP-AR)} \label{sec:gpar}

The simplest auto-regressive models for \emph{OSA} prediction of $Dst$ are those that use only the history of $Dst$ to construct input features for model training. The input features $\mathbf{x}_{t-1}$ at each time step are the history of $Dst(t)$ until a time lag of $p$ hours.

\begin{align*}
    \mathbf{x}_{t-1} & = & \left(Dst(t-1), \cdots , Dst(t-p+1)\right)
\end{align*}

\subsection{Gaussian Process Auto-Regressive with eXogenous inputs (GP-ARX)} \label{sec:gparx}

Auto-regressive models can be augmented by including exogenous quantities in the inputs $\mathbf{x}_{t-1}$ at each time step, in order to improve predictive accuracy. $Dst$ gives a measure of ring currents, which are modulated by plasma sheet particle injections into the inner magnetosphere during sub-storms. Studies have shown that the substorm occurrence rate increases with solar wind velocity (high speed streams) \citep{Kissinger2011,Newell2016}. Prolonged southward interplanetary magnetic field (IMF) $z$-component ($B_z$) is needed for sub-storms to occur \citep{McPherron1986}. An increase in the solar wind electric field, $VB_z$, can increase the dawn-dusk electric field in the magnetotail, which in turn determines the amount of plasma sheet particle that move to the inner magnetosphere \citep{Friedel2001}. Therefore, our exogenous parameters consist of solar wind velocity, IMF $B_z$, and $VB_z$.

In this model we choose distinct time lags $p$ and $p_{ex}$ for auto-regressive and exogenous variables respectively. In addition we explicitly include the product $V B_z$ which is a proxy for the electric field, as an input.

\begin{align*}
        & \mathbf{x}_{t-1} & = (Dst(t-1), \cdots , Dst(t-p+1), \\
        & \ \ \ \ \ & V(t-1), \cdots, V(t-p_{ex}+1),\\
        & \ \ \ \ \ & B_{z}(t-1), \cdots, B_{z}(t-p_{ex}+1),\\
        & \ \ \ \ \ & V B_{z}(t-1), \cdots, V B_{z}(t-p_{ex}+1))
\end{align*}

\section{Model Training and Validation}

Before running performance bench marks for \emph{OSA} $Dst$ prediction on the storm events in \citet{Ji2012}, training and model selection of \emph{GP-AR} and \emph{GP-ARX} models on independent data sets must be performed. For this purpose we choose segments 00:00 1 January 2008 - 10:00 11 January 2008 for training and 00:00 15 November 2014 - 23:00 1 December 2014 for model selection.

Although the training and model selection data sets both do not have a geomagnetic storm in them, this would not degrade the performance of \emph{GP-AR} and \emph{GP-ARX} because the linear polynomial kernel describes a non stationary and self similar Gaussian Process. This implies that for two events where the time histories of $Dst$, $V$ and $B_z$ are not close to each other but can be expressed as a diagonal rescaling of time histories observed in the training data, the predictive distribution is a linearly rescaled version of the training data $Dst$ distribution.

The computational complexity of calculation of the predictive distribution is $O(N^3)$, as discussed in section \ref{sec:inference}. This can limit the size of the covariance matrix constructed from the training data. Note that this computation overhead is paid for every unique assignment to the model hyper-parameters. However, our chosen training set has a size of 250 which is still very much below the computational limits of the method and in our case. Solving equation \ref{eq:posteriormean} on a laptop computer takes less than a second for the training set considered in our analysis.

The size of the training set has been decided by studying the performance of the models for increasing training sets. We have noticed that using training sets with more than 250 values does not increase the overall performance in \emph{OSA} prediction of Dst.

%%% End of body of article:

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Optional Appendix goes here
%
% \appendix resets counters and redefines section heads
% but doesn't print anything.
% After typing \appendix
%
%\section{Here Is Appendix Title}
% will show
% Appendix A: Here Is Appendix Title
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Optional Glossary or Notation section, goes here
%

%%%%%%%%%%%%%%
% Notation -- End each entry with a period.
% \begin{notation}
% Term & definition.\\
% Second term & second definition.\\
% \end{notation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  ACKNOWLEDGMENTS

\begin{acknowledgments}
We acknowledge use of NASA/GSFC's Space Physics Data Facility's OMNIWeb (or CDAWeb or ftp) service, and OMNI data. We also acknowledge the authors of \citet{JGRA:JGRA16300} for their scientific inputs towards understanding the \emph{TL} model and assistance in access of its generated predictions. Simon Wing acknowledges supports from CWI and NSF Grant AGS-1058456 and NASA Grants (NNX13AE12G, NNX15AJ01G, NNX16AC39G).
\end{acknowledgments}

%% ------------------------------------------------------------------------ %%
%%  REFERENCE LIST AND TEXT CITATIONS
%
% Either type in your references using
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% Or,
%
% If you use BiBTeX for your references, please use the agufull08.bst file (available at % ftp://ftp.agu.org/journals/latex/journals/Manuscript-Preparation/) to produce your .bbl
% file and copy the contents into your paper here.
%
% Follow these steps:
% 1. Run LaTeX on your LaTeX file.
%
% 2. Make sure the bibliography style appears as \bibliographystyle{agufull08}. Run BiBTeX on your LaTeX
% file.
%
% 3. Open the new .bbl file containing the reference list and
%   copy all the contents into your LaTeX file here.
%
% 4. Comment out the old \bibliographystyle and \bibliography commands.
%
% 5. Run LaTeX on your new file before submitting.
%
% AGU does not want a .bib or a .bbl file. Please copy in the contents of your .bbl file here.
\bibliographystyle{BibTeX/agufull08}
\bibliography{BibTeX/sample-bib-08}
%\begin{thebibliography}{}

%\providecommand{\natexlab}[1]{#1}
%\expandafter\ifx\csname urlstyle\endcsname\relax
%  \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
%  \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
%  \urlstyle{rm}\Url}\fi
%
%\bibitem[{\textit{Atkinson and Sloan}(1991)}]{AtkinsonSloan}
%Atkinson, K., and I.~Sloan (1991), The numerical solution of first-kind
%  logarithmic-kernel integral equations on smooth open arcs, \textit{Math.
%  Comp.}, \textit{56}(193), 119--139.
%
%\bibitem[{\textit{Colton and Kress}(1983)}]{ColtonKress1}
%Colton, D., and R.~Kress (1983), \textit{Integral Equation Methods in
%  Scattering Theory}, John Wiley, New York.
%
%\bibitem[{\textit{Hsiao et~al.}(1991)\textit{Hsiao, Stephan, and
%  Wendland}}]{StephanHsiao}
%Hsiao, G.~C., E.~P. Stephan, and W.~L. Wendland (1991), On the {D}irichlet
%  problem in elasticity for a domain exterior to an arc, \textit{J. Comput.
%  Appl. Math.}, \textit{34}(1), 1--19.
%
%\bibitem[{\textit{Lu and Ando}(2012)}]{LuAndo}
%Lu, P., and M.~Ando (2012), Difference of scattering geometrical optics
%  components and line integrals of currents in modified edge representation,
%  \textit{Radio Sci.}, \textit{47},  RS3007, \doi{10.1029/2011RS004899}.

%\end{thebibliography}

%Reference citation examples:

%...as shown by \textit{Kilby} [2008].
%...as shown by {\textit  {Lewin}} [1976], {\textit  {Carson}} [1986], {\textit  {Bartholdy and Billi}} [2002], and {\textit  {Rinaldi}} [2003].
%...has been shown [\textit{Kilby et al.}, 2008].
%...has been shown [{\textit  {Lewin}}, 1976; {\textit  {Carson}}, 1986; {\textit  {Bartholdy and Billi}}, 2002; {\textit  {Rinaldi}}, 2003].
%...has been shown [e.g., {\textit  {Lewin}}, 1976; {\textit  {Carson}}, 1986; {\textit  {Bartholdy and Billi}}, 2002; {\textit  {Rinaldi}}, 2003].

%...as shown by \citet{jskilby}.
%...as shown by \citet{lewin76}, \citet{carson86}, \citet{bartoldy02}, and \citet{rinaldi03}.
%...has been shown \citep{jskilbye}.
%...has been shown \citep{lewin76,carson86,bartoldy02,rinaldi03}.
%...has been shown \citep [e.g.,][]{lewin76,carson86,bartoldy02,rinaldi03}.
%
% Please use ONLY \citet and \citep for reference citations.
% DO NOT use other cite commands (e.g., \cite, \citeyear, \nocite, \citealp, etc.).

%% ------------------------------------------------------------------------ %%
%
%  END ARTICLE
%
%% ------------------------------------------------------------------------ %%
\end{article}
%
%
%% Enter Figures and Tables here:
%
% DO NOT USE \psfrag or \subfigure commands.
%
% Figure captions go below the figure.
% Table titles go above tables; all other caption information
%  should be placed in footnotes below the table.
%
%----------------
% EXAMPLE FIGURE
%
% \begin{figure}
% \noindent\includegraphics[width=20pc]{samplefigure.eps}
% \caption{Caption text here}
% \label{figure_label}
% \end{figure}

%
% ---------------
% EXAMPLE TABLE
%
%\begin{table}
%\caption{Time of the Transition Between Phase 1 and Phase 2\tablenotemark{a}}
%\centering
%\begin{tabular}{l c}
%\hline
% Run  & Time (min)  \\
%\hline
%  $l1$  & 260   \\
%  $l2$  & 300   \\
%  $l3$  & 340   \\
%  $h1$  & 270   \\
%  $h2$  & 250   \\
%  $h3$  & 380   \\
%  $r1$  & 370   \\
%  $r2$  & 390   \\
%\hline
%\end{tabular}
%\tablenotetext{a}{Footnote text here.}
%\end{table}

% See below for how to make sideways figures or tables.

\end{document}

% IF YOU HAVE MULTI-LINE EQUATIONS, PLEASE
% BREAK THE EQUATIONS INTO TWO OR MORE LINES
% OF SINGLE COLUMN WIDTH (20 pc, 8.3 cm)
% using double backslashes (\\).

% To create multiline equations, use the
% \begin{eqnarray} and \end{eqnarray} environment
% as demonstrated below.
\begin{eqnarray}
  x_{1} & = & (x - x_{0}) \cos \Theta \nonumber \\
        && + (y - y_{0}) \sin \Theta  \nonumber \\
  y_{1} & = & -(x - x_{0}) \sin \Theta \nonumber \\
        && + (y - y_{0}) \cos \Theta.
\end{eqnarray}

%If you don't want an equation number, use the star form:
%\begin{eqnarray*}...\end{eqnarray*}

% Break each line at a sign of operation
% (+, -, etc.) if possible, with the sign of operation
% on the new line.

% Indent second and subsequent lines to align with
% the first character following the equal sign on the
% first line.

% Use an \hspace{} command to insert horizontal space
% into your equation if necessary. Place an appropriate
% unit of measure between the curly braces, e.g.
% \hspace{1in}; you may have to experiment to achieve
% the correct amount of space.

%% ------------------------------------------------------------------------ %%
%
%  EQUATION NUMBERING: COUNTER
%
%% ------------------------------------------------------------------------ %%

% You may change equation numbering by resetting
% the equation counter or by explicitly numbering
% an equation.

% To explicitly number an equation, type \eqnum{}
% (with the desired number between the brackets)
% after the \begin{equation} or \begin{eqnarray}
% command.  The \eqnum{} command will affect only
% the equation it appears with; LaTeX will number
% any equations appearing later in the manuscript
% according to the equation counter.
%

% If you have a multiline equation that needs only
% one equation number, use a \nonumber command in
% front of the double backslashes (\\) as shown in
% the multiline equation above.

%% ------------------------------------------------------------------------ %%
%
%  SIDEWAYS FIGURE AND TABLE EXAMPLES
%
%% ------------------------------------------------------------------------ %%
%
% For tables and figures, add \usepackage{rotating} to the paper and add the rotating.sty file to the folder.
% AGU prefers the use of {sidewaystable} over {landscapetable} as it causes fewer problems.
%
% \begin{sidewaysfigure}
% \includegraphics[width=20pc]{samplefigure.eps}
% \caption{caption here}
% \label{label_here}
% \end{sidewaysfigure}
%
%
%
% \begin{sidewaystable}
% \caption{}
% \begin{tabular}
% Table layout here.
% \end{tabular}
% \end{sidewaystable}
%
%

### 2017-02-13 agutmpl.tex
Added corresponding author
Added abstract and partial introduction

@@
 %Corresponding author mailing address and e-mail address:
-%\authoraddr{Corresponding author: A. B. Smith,
-%Department of Hydrology and Water Resources, University of
-%Arizona, Harshbarger Building 11, Tucson, AZ 85721, USA.
-%(a.b.smith@hwr.arizona.edu)}
+\authoraddr{Corresponding author: M. H. Chandorkar,
+Multiscale Dynamics, Centrum Wiskunde Informatica, Science Park 123, 1098XG Amsterdam, Netherlands.
+(mandar.chandorkar@cwi.nl)}
 
@@
 \begin{abstract}
-(Type abstract here)
+We present a methodology for generating probabilistic predictions for the \emph{Disturbance Storm Time} ($Dst$) geomagnetic activity index.
+We focus on the \emph{One Step Ahead} (OSA) prediction task and use the OMNI hourly resolution data to build our models.
+
+Our proposed methodology is based on the technique of \emph{Gaussian Process Regression} (GPR).
+Within this framework we develop two models; \emph{Gaussian Process Auto-Regressive} (GP-AR) and \emph{Gaussian Process Auto-Regressive with eXogenous inputs} (GP-ARX).
+
+We also propose a criterion to aid model selection with respect to the order of auto-regressive inputs.
+Finally we test the performance of the GP-AR and GP-ARX models on a set of 63 geomagnetic storms between 1998 and 2006 and illustrate sample predictions with error bars for some of these events.
+
 \end{abstract}
@@
 \section{Introduction}
-(Article text here)
+
+The magnetosphere's dynamics and its associated solar wind driver form a complex dynamical system.
+It is therefore instructive and greatly simplifying to use representative indices to quantify the state of geomagnetic activity.
+
+Geomagnetic indices come in various forms, they may take continuous or discrete values and may be defined with varying time resolutions.
+Their values are often calculated by averaging or combining a number of readings taken by instruments around the Earth.
+Each geomagnetic index is a proxy for a particular kind of phenomenon.
+Some popular indices are the $K_p$, $Dst$ and the $AE$ index.
+
+\begin{enumerate}
+    \item $K_p$: The Kp-index is a discrete valued global geomagnetic storm index and is based on 3 hour measurements of the K-indices \citep{Bartels}.
+The K-index itself is a three hour long quasi-logarithmic local index of the geomagnetic activity, relative to a calm day curve for the given location.
+
+    \item $AE$: The Auroral Electrojet Index, $AE$, is designed to provide a global, quantitative measure of auroral zone magnetic activity produced by enhanced Ionospheric currents flowing below and within the auroral oval \citep{AEIndex}.
+It is a continuous index which is calculated every hour.
+
+    \item $Dst$: A continuous hourly index which measures the weakening of the Earths magnetic field due to ring currents and the strength of geomagnetic storms \citep{DesslerAndParker}. 
+\end{enumerate}
+
+%Talk about Burton and friends
+For the present study, we focus on prediction of the hourly $Dst$ index which is a straightforward indicator of geomagnetic storms.
+More specifically, we focus on the \emph{one step ahead} (OSA), in this case one hour ahead prediction of $Dst$ because it is the simplest model towards building long term predictions of geomagnetic response of the Earth to changing space weather conditions.
+
+The $Dst$ OSA prediction problem has been the subject of several modeling efforts in the literature.
+One of the earliest models has been presented by \citet{JGR:JGR10260} who calculated $Dst(t)$ as the solution of an \emph{Ordinary Differential Equation} (ODE) which expressed the rate of change of $Dst(t)$ as a combination of two terms: decay and injection $\frac{d Dst(t)}{dt} = Q(t) - \frac{Dst(t)}{\tau}$, where Q(t) relates to the particle injection from the plasma sheet into the inner magnetosphere.
+
+The \citet{JGR:JGR10260} model has proven to be very influential particularly due to its simplicity.
+Many subsequent works have modified the proposed ODE by proposing alternative expressions for the injection term $Q(t)$ [see \citet{Wang:Dst}, \citet{JGRA:JGRA14856}].
+More recently \citet{Ballatore2014} have tried to generate empirical estimates for the injection and decay terms in Burton's equation.
+
+%Talk about NARMAX Dst
+Another important empirical model used to predict $Dst$ is the \emph{Nonlinear Auto-Regessive Moving Average with eXogenous inputs} (NARMAX) methodology developed in \citet{doi:10.1080/00207178908559767}, \citet{GRL:GRL13494}, \citet{GRL:GRL20944}, \citet{JGRA:JGRA18657}, \citet{balikhin:narmax}, \citet{JGRA:JGRA20661} and \citet{JGRA:JGRA50192}.
+The NARMAX methodology builds models by constructing polynomial expansions of inputs and determines the best combinations of monomials to include in the refined model by using a criterion called the \emph{error reduction ratio} (ERR).
+The parameters of the so called NARMAX OLS-ERR model are calculated by solving the \emph{ordinary least squares} (OLS) problem arising from a quadratic objective objective function.
+The reader may refer to \citet{billings2013nonlinear} for a detailed exposition of the NARMAX methodology.
+
+%Talk about neural networks
+Yet another family of forecasting methods is based on \emph{Artificial Neural Networks} (ANN) that have been a popular choice for building predictive models.
+Researchers have employed both the standard \emph{feed forward} and the more specialized \emph{recurrent} architectures.
+\citet{Lund} proposed an \emph{Elman} recurrent network architecture called Lund $Dst$, which used the solar wind velocity, \emph{interplanetary magnetic field} (IMF) and historical $Dst$ data as inputs.
+\citet{JGRA:JGRA17461} used recurrent neural networks to predict $Kp$.
+\citet{SWE:SWE286} originally proposed a \emph{feed forward} network for predicting the $K_p$ index which used the \emph{Boyle coupling function} \citep{boyle1997empirical}.
+The same architecture is adapted for prediction of $Dst$ in \citet{SWE:SWE286}, popularly known as the Rice $Dst$ model.
+\citet{pallocchia:hal-00318011} proposed a \emph{neural network} model called EDDA to predict $Dst$ using only the IMF data.
 

### 2017-02-14 agutmpl.tex
Added figures

@@
 \begin{align}
-  \mathbf{X} & = & \left( \begin{array}{c} \mathbf{x}^{T}_1 \\ \mathbf{x}^{T}_2 \\ \vdots \\ \mathbf{x}^{T}_n \end{array} \right)_{n \times d} \label{eq:feat} \\
+  \mathbf{X}  = & \left( \begin{array}{c} \mathbf{x}^{T}_1 \\ \mathbf{x}^{T}_2 \\ \vdots \\ \mathbf{x}^{T}_n \end{array} \right)_{n \times d} \label{eq:feat} \\
   \vspace{2\baselineskip}
-  \mathbf{y} & = & \left( \begin{array}{c} y_1 \\ y_2 \\ \vdots \\ y_N \end{array} \right) _{n \times 1} \label{eq:labels}
+  \mathbf{y}  = & \left( \begin{array}{c} y_1 \\ y_2 \\ \vdots \\ y_N \end{array} \right) _{n \times 1} \label{eq:labels}
 \end{align}
@@
 \begin{align}
- \mathbf{f} & = & \left( \begin{array}{c} f(\mathbf{x}_1) \\ f(\mathbf{x}_2) \\ \vdots \\ f(\mathbf{x}_N) \end{array} \right) \label{eq:fvalues}\\
+ \mathbf{f} = & \left( \begin{array}{c} f(\mathbf{x}_1) \\ f(\mathbf{x}_2) \\ \vdots \\ f(\mathbf{x}_N) \end{array} \right) \label{eq:fvalues}\\
  \vspace{2\baselineskip}
- \mathbf{f} | \mathbf{x}_1, \cdots, \mathbf{x}_N & \sim & \mathcal{N}\left( \mathbf{\mu}, \mathbf{\Lambda} \right)  \label{eq:normal}\\
+ \mathbf{f} | \mathbf{x}_1, \cdots, \mathbf{x}_N \sim & \mathcal{N}\left( \mathbf{\mu}, \mathbf{\Lambda} \right)  \label{eq:normal}\\
  \vspace{2\baselineskip}
- p( \mathbf{f} \ | \ \mathbf{x}_1, \cdots, \mathbf{x}_N) & = & \frac{1}{(2\pi)^{n/2} det(\mathbf{\Lambda})^{1/2}} exp \left(-\frac{1}{2} (\mathbf{f} - \mathbf{\mu})^T \mathbf{\Lambda}^{-1} (\mathbf{f} - \mathbf{\mu}) \right) \label{eq:sto}
+ p( \mathbf{f} \ | \ \mathbf{x}_1, \cdots, \mathbf{x}_N) = & \frac{1}{(2\pi)^{n/2} det(\mathbf{\Lambda})^{1/2}} exp \left(-\frac{1}{2} (\mathbf{f} - \mathbf{\mu})^T \mathbf{\Lambda}^{-1} (\mathbf{f} - \mathbf{\mu}) \right) \label{eq:sto}
 \end{align}
@@
 \begin{align}
-      \mu_i & = & \mathbb{E}[f(\mathbf{x}_i)] := m(\mathbf{x}_i) \\
-      \Lambda_{ij} & = & \mathbb{E}[(f(\mathbf{x}_i) - \mu_i)(f(\mathbf{x}_j) - \mu_j)] := K(\mathbf{x}_i, \mathbf{x}_j)
+      \mu_i = & \mathbb{E}[f(\mathbf{x}_i)] := m(\mathbf{x}_i) \\
+      \Lambda_{ij} = & \mathbb{E}[(f(\mathbf{x}_i) - \mu_i)(f(\mathbf{x}_j) - \mu_j)] := K(\mathbf{x}_i, \mathbf{x}_j)
 \end{align}
@@
 \begin{align}
-    \mathbf{f}_* & = & \left( \begin{array}{c} f(\mathbf{x^{*}_1}) \\ f(\mathbf{x^{*}_2}) \\ \vdots \\ f(\mathbf{x^{*}_M}) \end{array} \right)_{M \times 1} \\
+    \mathbf{f}_* = & \left( \begin{array}{c} f(\mathbf{x^{*}_1}) \\ f(\mathbf{x^{*}_2}) \\ \vdots \\ f(\mathbf{x^{*}_M}) \end{array} \right)_{M \times 1} \\
      \vspace{4\baselineskip}
-    \left( \begin{array}{c} \mathbf{y} \\ \mathbf{f_*} \end{array} \right) | \ \ \mathbf{X}, \mathbf{X}_* & \sim & 
+    \left( \begin{array}{c} \mathbf{y} \\ \mathbf{f_*} \end{array} \right) | \ \ \mathbf{X}, \mathbf{X}_* \sim & 
     \mathcal{N}\left(\mathbf{0}, \left[ \begin{array}{cc} \mathbf{K} + \sigma^{2} \mathbf{I} & \mathbf{K}_{*} \\ \mathbf{K}_{*}^T & \mathbf{K}_{**} \end{array} \right ] \right) \label{eq:dist}
@@
 \begin{align}
-    \mathbf{\bar{f}_*} & = & \mathbf{K}^T_{*} [\mathbf{K} + \sigma^{2} \mathbf{I}]^{-1} \mathbf{y} \label{eq:posteriormean} \\
-    \Sigma_* & = & \mathbf{K}_{**} - \mathbf{K}^T_{*} \left(\mathbf{K} + \sigma^{2} \mathbf{I}\right)^{-1} \mathbf{K}_{*} \label{eq:posteriorcov}
+    \mathbf{\bar{f}_*} = & \mathbf{K}^T_{*} [\mathbf{K} + \sigma^{2} \mathbf{I}]^{-1} \mathbf{y} \label{eq:posteriormean} \\
+    \Sigma_* = & \mathbf{K}_{**} - \mathbf{K}^T_{*} \left(\mathbf{K} + \sigma^{2} \mathbf{I}\right)^{-1} \mathbf{K}_{*} \label{eq:posteriorcov}
 \end{align}
@@
 
-\begin{equation}\label{eq:usedKernel}
-    K(\mathbf{x}, \mathbf{y}) = \mathbf{x}^\intercal \mathbf{y}
-\end{equation}
+\begin{align}
+    K_{osa}(\mathbf{x}, \mathbf{y}) & = K_{mlp}(\mathbf{x}, \mathbf{y}) + K_{st}(\mathbf{x}, \mathbf{y}) \label{eq:usedKernel} \\
+    K_{mlp}(\mathbf{x}, \mathbf{y}) & = sin^{-1}(\frac{w\mathbf{x}^\intercal \mathbf{y} + b}{\sqrt{w\mathbf{x}^\intercal \mathbf{x} + b + 1} \sqrt{w\mathbf{y}^\intercal \mathbf{y} + b + 1}}) \\
+    K_{st}(\mathbf{x}, \mathbf{y}) & = \frac{1}{1 + ||\mathbf{x} - \mathbf{y}||_{2}^d}
+\end{align}
 
@@
 \begin{align}
-    Dst(t) & = & f(\mathbf{x}_{t-1}) + \epsilon \label{eq:Dst} \\
-    \epsilon & \sim & \mathcal{N}(0, \sigma^2) \label{eq:GPNoise} \\
-    f(x_t) & \sim & \mathcal{GP}(m(\mathbf{x}_t), K(\mathbf{x}_t, \mathbf{x}_s)) \label{eq:DstGP} \\
-    K(\mathbf{x}, \mathbf{y}) & = & \mathbf{x}^\intercal \mathbf{y} + b \label{eq:GPPoly}
+    Dst(t) & =  f(\mathbf{x}_{t-1}) + \epsilon \label{eq:Dst} \\
+    \epsilon & \sim  \mathcal{N}(0, \sigma^2) \label{eq:GPNoise} \\
+    f(x_t) & \sim  \mathcal{GP}(m(\mathbf{x}_t), K(\mathbf{x}_t, \mathbf{x}_s)) \label{eq:DstGP} \\
+    K(\mathbf{x}, \mathbf{y}) & =  K_{osa}(\mathbf{x}, \mathbf{y}) \label{eq:GPPoly}
 \end{align}
@@
 \begin{align*}
-    \mathbf{x}_{t-1} & = & \left(Dst(t-1), \cdots , Dst(t-p+1)\right)
+    \mathbf{x}_{t-1} = & \left(Dst(t-1), \cdots , Dst(t-p+1)\right)
 \end{align*}
@@
 
-In this model we choose distinct time lags $p$ and $p_{ex}$ for auto-regressive and exogenous variables respectively.
-In addition we explicitly include the product $V B_z$ which is a proxy for the electric field, as an input.
+In this model we choose distinct time lags $p$, $p_{v}$ and $p_{b}$ for $Dst$, $V$ and $B_z$ respectively.
 
 \begin{align*}
-        & \mathbf{x}_{t-1} & = (Dst(t-1), \cdots , Dst(t-p+1), \\
-        & \ \ \ \ \ & V(t-1), \cdots, V(t-p_{ex}+1),\\
-        & \ \ \ \ \ & B_{z}(t-1), \cdots, B_{z}(t-p_{ex}+1),\\
-        & \ \ \ \ \ & V B_{z}(t-1), \cdots, V B_{z}(t-p_{ex}+1))
+       \mathbf{x}_{t-1} & = (Dst(t-1), \cdots , Dst(t-p+1), \\
+        & \ \ \ \ \  V(t-1), \cdots, V(t-p_{v}+1),\\
+        & \ \ \ \ \  B_{z}(t-1), \cdots, B_{z}(t-p_{b}+1))
 \end{align*}

### 2017-02-14 agutmpl.tex
Added references

@@
 
-\title{Gaussian Processes Autoregressive Models for Forecasting the Disturbance Storm Time Index}
+\title{Probabilistic Forecasting of the Disturbance Storm Time Index: An Autoregressive Gaussian Process approach}
 
@@
 
+%Talk about need for probabilistic forecasts.
+Although much research has been done on prediction of the $Dst$ index, must less has been done on probabilistic forecasting of $Dst$.
+One such work described in \citet{McPherron:2013} involves identification of high speed solar wind streams using the W-S-A model (see \citet{WSAModel}), using predictions of high speed streams to construct ensembles of $Dst$ trajectories which yield the quartiles of $Dst$ time series.
+
+In this work we propose an inductive model for probabilistic forecating of the OSA $Dst$ using only the OMNI data.
+We use the \emph{Gaussian Process Regression} methodology to construct auto-regressive models for $Dst$ and show how to perform exact inference in this framework.
+We further outline a methodology to perform model selection with respect to its free parameters and time histories.
+
 \section{Methodology: Gaussian Process} \label{sec:method}
@@
 \end{equation}
+
 Using the multivariate Gaussian distribution in equation (\ref{eq:sto}) we can construct the joint distribution of $f(\mathbf{x})$ over the training and test points.

### 2017-02-15 agutmpl.tex
Added figures

@@
 \usepackage{txfonts}
+\usepackage{graphicx}
 \usepackage{url}
@@
 
+\graphicspath{{./figures/}}
 %  To add line numbers to lines with equations:
@@
 
-In this work we propose an inductive model for probabilistic forecating of the OSA $Dst$ using only the OMNI data.
+In this work we propose an technique for probabilistic forecating of $Dst$, which yields a predictive distribution as a closed form expression.
+Our models take as input past values of $Dst$, solar wind speed and the z-component of the \emph{Interplanetary Magnetic Field} (IMF) and output a Gaussian distribution with a specific mean and variance as the OSA prediction of the $Dst$.
+
 We use the \emph{Gaussian Process Regression} methodology to construct auto-regressive models for $Dst$ and show how to perform exact inference in this framework.
 We further outline a methodology to perform model selection with respect to its free parameters and time histories.
+
+The remainder of this paper is organised as follows: Section \ref{sec:method} gives the reader an overview of the history of \emph{Gaussian Process} models as well as how they are formulated and how to perform inference with them.
+Sections \ref{sec:osa}, \ref{sec:modeltraining} describe the GP-AR and GP-ARX models for OSA prediction of $Dst$ and how to choose their free parameters for better performance.
 
@@
 Since the original distribution of $\left( \begin{array}{c} \mathbf{y} \\ \mathbf{f_*} \end{array} \right) | \ \ \mathbf{X}, \mathbf{X}_*$ is a multivariate Gaussian, conditioning on a subset of elements $\mathbf{y}$ yields another Gaussian distribution whose mean and covariance can be calculated exactly, as in equation (\ref{eq:posterior}) (see \citet{Rasmussen:2005:GPM:1162254}).
+
 \begin{equation}
@@
 The interested reader is suggested to refer to \citet{Berlinet2004}, \citet{Scholkopf:2001:LKS:559923} and \citet{hofmann2008} for a thorough treatment of kernel functions and the rich theory behind them.
-Some common kernel functions used in machine learning include the radial basis function (RBF) kernel $K(\mathbf{x}, \mathbf{y}) = \frac{1}{2} exp(-||\mathbf{x} - \mathbf{y}||^2/l^2)$ and the polynomial kernel $K(\mathbf{x}, \mathbf{y}) = (\mathbf{x}^\intercal \mathbf{y} + b)^d$.
+Some common kernel functions used in machine learning are listed in table \ref{table:kernel}.
 
@@
 Hyper-parameters give flexibility to particular kernel structure, for example $d = 1, 2, 3, \cdots$ in the polynomial kernel represents linear, quadratic, cubic and higher order polynomials respectively.
-The method of assigning values to the \emph{hyper-parameters} is crucial in the model building process.
+The method of assigning values to the \emph{hyper-parameters} is crucial in the model building process and is known as \emph{model selection}.
 
-In this study, we construct Gaussian Process regression models with linear kernels as shown in equation (\ref{eq:usedKernel}), for \emph{one step ahead} prediction of the $Dst$ index.
-Our choice of kernel leaves us with only one adjustable hyper-parameter, the model noise $\sigma$.
-
-\begin{align}
-    K_{osa}(\mathbf{x}, \mathbf{y}) & = K_{mlp}(\mathbf{x}, \mathbf{y}) + K_{st}(\mathbf{x}, \mathbf{y}) \label{eq:usedKernel} \\
-    K_{mlp}(\mathbf{x}, \mathbf{y}) & = sin^{-1}(\frac{w\mathbf{x}^\intercal \mathbf{y} + b}{\sqrt{w\mathbf{x}^\intercal \mathbf{x} + b + 1} \sqrt{w\mathbf{y}^\intercal \mathbf{y} + b + 1}}) \\
-    K_{st}(\mathbf{x}, \mathbf{y}) & = \frac{1}{1 + ||\mathbf{x} - \mathbf{y}||_{2}^d}
-\end{align}
-
-We initialize a grid of values for the model noise $\sigma$ and use the error on a predefined validation data set to choose the best performing value of $\sigma$.
-While constructing the grid of possible values for $\sigma$ it must be ensured that $\sigma > 0$ so that the kernel matrix constructed on the training data is non-singular.
+\subsection{Model Selection}
 
@@
     K(\mathbf{x}, \mathbf{y}) & =  K_{osa}(\mathbf{x}, \mathbf{y}) \label{eq:GPPoly}
+\end{align}
+
+\subsection{Choice of Kernel}
+
+In this study, we construct Gaussian Process regression models with a combination of the \emph{maximum likelihood perceptron} kernel and \emph{student's T} kernel as shown in equations (\ref{eq:usedKernel}), for \emph{one step ahead} prediction of the $Dst$ index.
+
+\begin{align}
+    K_{osa}(\mathbf{x}, \mathbf{y}) & = K_{mlp}(\mathbf{x}, \mathbf{y}) + K_{st}(\mathbf{x}, \mathbf{y}) \label{eq:usedKernel} \\
+    K_{mlp}(\mathbf{x}, \mathbf{y}) & = sin^{-1}(\frac{w\mathbf{x}^\intercal \mathbf{y} + b}{\sqrt{w\mathbf{x}^\intercal \mathbf{x} + b + 1} \sqrt{w\mathbf{y}^\intercal \mathbf{y} + b + 1}}) \\
+    K_{st}(\mathbf{x}, \mathbf{y}) & = \frac{1}{1 + ||\mathbf{x} - \mathbf{y}||_{2}^d}
 \end{align}
@@
 An increase in the solar wind electric field, $VB_z$, can increase the dawn-dusk electric field in the magnetotail, which in turn determines the amount of plasma sheet particle that move to the inner magnetosphere \citep{Friedel2001}.
-Therefore, our exogenous parameters consist of solar wind velocity, IMF $B_z$, and $VB_z$.
+Therefore, our exogenous parameters consist of solar wind velocity and IMF $B_z$.
 
@@
 
-\section{Model Training and Validation}
+\section{Model Training and Validation} \label{sec:modeltraining}
 
@@
 For this purpose we choose segments 00:00 1 January 2008 - 10:00 11 January 2008 for training and 00:00 15 November 2014 - 23:00 1 December 2014 for model selection.
-
-Although the training and model selection data sets both do not have a geomagnetic storm in them, this would not degrade the performance of \emph{GP-AR} and \emph{GP-ARX} because the linear polynomial kernel describes a non stationary and self similar Gaussian Process.
-This implies that for two events where the time histories of $Dst$, $V$ and $B_z$ are not close to each other but can be expressed as a diagonal rescaling of time histories observed in the training data, the predictive distribution is a linearly rescaled version of the training data $Dst$ distribution.
 
@@
 
-The size of the training set has been decided by studying the performance of the models for increasing training sets.
-We have noticed that using training sets with more than 250 values does not increase the overall performance in \emph{OSA} prediction of Dst.
+\section{Experiments}\label{sec:exp}
+
+\section{Conclusions}\label{sec:res}
 
@@
 We acknowledge use of NASA/GSFC's Space Physics Data Facility's OMNIWeb (or CDAWeb or ftp) service, and OMNI data.
-We also acknowledge the authors of \citet{JGRA:JGRA16300} for their scientific inputs towards understanding the \emph{TL} model and assistance in access of its generated predictions.
 Simon Wing acknowledges supports from CWI and NSF Grant AGS-1058456 and NASA Grants (NNX13AE12G, NNX15AJ01G, NNX16AC39G).
@@
 
+\begin{figure}
+\noindent\includegraphics[width=20pc]{Compare-mae-arx.eps}
+\caption{Mean Absolute Error on validation set storms vs model order for GP-AR and GP-ARX}
+\label{fig:CompareMae}
+\end{figure}
+
 %
@@
 %\end{table}
+
+\begin{table}
+\caption{Popular Kernel functions used in GPR models}
+\centering
+\begin{tabular}{l c c}
+\hline
+ Name  & Expression & Hyperparameters  \\
+\hline
+  Radial Basis Function (RBF)  & $\frac{1}{2} exp(-||\mathbf{x} - \mathbf{y}||^2/l^2)$  & $l \in \mathbb{R}$   \\
+
+  Polynomial  & $(\mathbf{x}^\intercal \mathbf{y} + b)^d$ & $b \in \mathbb{R}, d \in \mathbb{N}$   \\
+
+  Laplacian  & $exp(-||\mathbf{x} - \mathbf{y}||_{1}/\theta)$  & $\theta \in \mathbb{R}^+$  \\
+
+  Student's T  & $1/(1 + ||\mathbf{x} - \mathbf{y}||_{2}^d)$ & $d \in \mathbb{R}^{+}$\\
+
+  Maximum Likelihood Perceptron  & $sin^{-1}(\frac{w\mathbf{x}^\intercal \mathbf{y} + b}{\sqrt{w\mathbf{x}^\intercal \mathbf{x} + b + 1} \sqrt{w\mathbf{y}^\intercal \mathbf{y} + b + 1}})$ & $w, b \in \mathbb{R}^{+}$\\
+\hline
+\end{tabular}
+\label{table:kernel}
+\end{table}
+
+% & $b \in \mathbb{R}, d \in \mathbb{N}$  & $l \in \mathbb{R}$
+%  & $d \in \mathbb{R}^{+}$
+%  & $w b \in \mathbb{R}^{+}$
 

### 2017-02-15 agutmpl.tex
Added model selection [WIP]

@@
 
-\documentclass[draft,sw]{AGUTeX}
+\documentclass[sw, draft]{AGUTeX}
 
@@
 \linenumbers*[1]
-
-\graphicspath{{./figures/}}
+\usepackage{rotating}
+\graphicspath{{figures/}}
 %  To add line numbers to lines with equations:
@@
 Hyper-parameters give flexibility to particular kernel structure, for example $d = 1, 2, 3, \cdots$ in the polynomial kernel represents linear, quadratic, cubic and higher order polynomials respectively.
-The method of assigning values to the \emph{hyper-parameters} is crucial in the model building process and is known as \emph{model selection}.
+The process of assigning values to the \emph{hyper-parameters} is crucial in the model building process and is known as \emph{model selection}.
 
@@
 
+Given a GP model with a kernel function $K_\theta$, the problem of model selection consists of finding appropriate values to the kernel hyper-parameters $\theta$.
+In order to assign values to $\theta$, we must define an objective function which represents our confidence that the GP model built from a particular value of $\theta$ is the best performing model.
+Since GP models encode assumptions about the probability distribution of the output data $\mathbf{y}$ given inputs $\mathbf{X}$, it is natural to use the negative log-likelihood of the training data as a model selection criterion.
+
+\begin{align*}
+  Q(\theta) & = - log(p(\mathbf{y}|\mathbf{X}, K_\theta)) \\
+            & = -\frac{1}{2} \mathbf{y}^\intercal (\mathbf{K}_\theta + \sigma^{2}_n \mathbf{I})^{-1} \mathbf{y} - \frac{1}{2}|\mathbf{K}_\theta + \sigma^{2}_n \mathbf{I}| - \frac{n}{2}log(2\pi) \\
+  \mathbf{K}_\theta & = [K_{\theta}(\mathbf{x}_i, \mathbf{x}_j)]_{n \times n}
+\end{align*}
+
 \section{One Step Ahead Prediction} \label{sec:osa}
 
-Below in equations (\ref{eq:Dst}) - (\ref{eq:GPPoly}) we outline a \emph{Gaussian Process} formulation for \emph{OSA} prediction of $Dst$.
+Below in equations (\ref{eq:Dst}) - (\ref{eq:DstGP}) we outline a \emph{Gaussian Process} formulation for \emph{OSA} prediction of $Dst$.
 A vector of features $\mathbf{x}_{t-1}$ is used as input to an unknown function $f(\mathbf{x}_{t-1})$.
@@
     \epsilon & \sim  \mathcal{N}(0, \sigma^2) \label{eq:GPNoise} \\
-    f(x_t) & \sim  \mathcal{GP}(m(\mathbf{x}_t), K(\mathbf{x}_t, \mathbf{x}_s)) \label{eq:DstGP} \\
-    K(\mathbf{x}, \mathbf{y}) & =  K_{osa}(\mathbf{x}, \mathbf{y}) \label{eq:GPPoly}
+    f(x_t) & \sim  \mathcal{GP}(m(\mathbf{x}_t), K_{osa}(\mathbf{x}_t, \mathbf{x}_s)) \label{eq:DstGP} \\
 \end{align}
 
-\subsection{Choice of Kernel}
-
-In this study, we construct Gaussian Process regression models with a combination of the \emph{maximum likelihood perceptron} kernel and \emph{student's T} kernel as shown in equations (\ref{eq:usedKernel}), for \emph{one step ahead} prediction of the $Dst$ index.
-
-\begin{align}
-    K_{osa}(\mathbf{x}, \mathbf{y}) & = K_{mlp}(\mathbf{x}, \mathbf{y}) + K_{st}(\mathbf{x}, \mathbf{y}) \label{eq:usedKernel} \\
-    K_{mlp}(\mathbf{x}, \mathbf{y}) & = sin^{-1}(\frac{w\mathbf{x}^\intercal \mathbf{y} + b}{\sqrt{w\mathbf{x}^\intercal \mathbf{x} + b + 1} \sqrt{w\mathbf{y}^\intercal \mathbf{y} + b + 1}}) \\
-    K_{st}(\mathbf{x}, \mathbf{y}) & = \frac{1}{1 + ||\mathbf{x} - \mathbf{y}||_{2}^d}
-\end{align}
-
-In the following we consider two choices for the input features $\mathbf{x}_{t-1}$ leading to two variants of \emph{Gaussian Process} regression for $Dst$ time series prediction.
+We consider two choices for the input features $\mathbf{x}_{t-1}$ leading to two variants of \emph{Gaussian Process} regression for $Dst$ time series prediction.
 
@@
 \end{align*}
+
+\subsection{Choice of Kernel}
+
+In this study, we construct Gaussian Process regression models with a combination of the \emph{maximum likelihood perceptron} kernel and \emph{student's T} kernel as shown in equations (\ref{eq:usedKernel}), for \emph{one step ahead} prediction of the $Dst$ index.
+
+\begin{align}
+    K_{osa}(\mathbf{x}, \mathbf{y}) & = K_{mlp}(\mathbf{x}, \mathbf{y}) + K_{st}(\mathbf{x}, \mathbf{y}) \label{eq:usedKernel} \\
+    K_{mlp}(\mathbf{x}, \mathbf{y}) & = sin^{-1}(\frac{w\mathbf{x}^\intercal \mathbf{y} + b}{\sqrt{w\mathbf{x}^\intercal \mathbf{x} + b + 1} \sqrt{w\mathbf{y}^\intercal \mathbf{y} + b + 1}}) \\
+    K_{st}(\mathbf{x}, \mathbf{y}) & = \frac{1}{1 + ||\mathbf{x} - \mathbf{y}||_{2}^d}
+\end{align}
 
@@
 \begin{figure}
-\noindent\includegraphics[width=20pc]{Compare-mae-arx.eps}
+\noindent\includegraphics[width=\textwidth]{Compare-mae.png}
 \caption{Mean Absolute Error on validation set storms vs model order for GP-AR and GP-ARX}
 \label{fig:CompareMae}
+\end{figure}
+
+\begin{figure}
+\noindent\includegraphics[width=\textwidth]{Compare-cc.png}
+\caption{Mean Absolute Error on validation set storms vs model order for GP-AR and GP-ARX}
+\label{fig:CompareCC}
+\end{figure}
+
+\begin{figure}
+\noindent\includegraphics[width=\textwidth]{Compare-mae-arx.png}
+\caption{Mean Absolute Error on validation set storms vs model order for GP-AR and GP-ARX}
+\label{fig:CompareMaeARX}
+\end{figure}
+
+\begin{figure}
+\noindent\includegraphics[width=\textwidth]{Compare-cc-arx.png}
+\caption{Mean Absolute Error on validation set storms vs model order for GP-AR and GP-ARX}
+\label{fig:CompareCCARX}
+\end{figure}
+
+\begin{figure}
+\noindent\includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm5.png}
+\caption{OSA Predictions with $\pm \sigma$ error bars for event: }
+\label{fig:ComparePred1}
+\end{figure}
+
+\begin{figure}
+\noindent\includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm8.png}
+\caption{OSA Predictions with $\pm \sigma$ error bars for event: }
+\label{fig:ComparePred2}
+\end{figure}
+
+\begin{figure}
+\noindent\includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm14.png}
+\caption{OSA Predictions with $\pm \sigma$ error bars for event: }
+\label{fig:ComparePred54}
 \end{figure}

### 2017-02-15 agutmpl.tex
Added section on mean function

@@
 \begin{enumerate}
-    \item $K_p$: The Kp-index is a discrete valued global geomagnetic storm index and is based on 3 hour measurements of the K-indices \citep{Bartels}.
+    \item $K_p$: The Kp-index is a discrete valued global geomagnetic storm index and is based on 3 hour measurements of the K-indices \citet{Bartels}.
 The K-index itself is a three hour long quasi-logarithmic local index of the geomagnetic activity, relative to a calm day curve for the given location.
 
-    \item $AE$: The Auroral Electrojet Index, $AE$, is designed to provide a global, quantitative measure of auroral zone magnetic activity produced by enhanced Ionospheric currents flowing below and within the auroral oval \citep{AEIndex}.
+    \item $AE$: The Auroral Electrojet Index, $AE$, is designed to provide a global, quantitative measure of auroral zone magnetic activity produced by enhanced Ionospheric currents flowing below and within the auroral oval \citet{AEIndex}.
 It is a continuous index which is calculated every hour.
 
-    \item $Dst$: A continuous hourly index which measures the weakening of the Earths magnetic field due to ring currents and the strength of geomagnetic storms \citep{DesslerAndParker}. 
+    \item $Dst$: A continuous hourly index which measures the weakening of the Earths magnetic field due to ring currents and the strength of geomagnetic storms \citet{DesslerAndParker}. 
 \end{enumerate}
@@
 \citet{JGRA:JGRA17461} used recurrent neural networks to predict $Kp$.
-\citet{SWE:SWE286} originally proposed a \emph{feed forward} network for predicting the $K_p$ index which used the \emph{Boyle coupling function} \citep{boyle1997empirical}.
+\citet{SWE:SWE286} originally proposed a \emph{feed forward} network for predicting the $K_p$ index which used the \emph{Boyle coupling function} \citet{boyle1997empirical}.
 The same architecture is adapted for prediction of $Dst$ in \citet{SWE:SWE286}, popularly known as the Rice $Dst$ model.
@@
 
-Given a GP model with a kernel function $K_\theta$, the problem of model selection consists of finding appropriate values to the kernel hyper-parameters $\theta$.
-In order to assign values to $\theta$, we must define an objective function which represents our confidence that the GP model built from a particular value of $\theta$ is the best performing model.
+Given a GP model with a kernel function $K_\theta$, the problem of model selection consists of finding appropriate values for the kernel hyper-parameters $\theta = \left(\theta_1, \theta_2, \cdots, \theta_i\right)$.
+In order to assign a value to $\theta$, we must define an objective function which represents our confidence that the GP model built from a particular value of $\theta$ is the best performing model.
 Since GP models encode assumptions about the probability distribution of the output data $\mathbf{y}$ given inputs $\mathbf{X}$, it is natural to use the negative log-likelihood of the training data as a model selection criterion.
@@
 \end{align*}
+
+The model selection problem can now be expressed as the minimization problem shown below.
+
+\begin{align*}
+\theta^* = arg\min_{\theta} \ Q(\theta)
+\end{align*}
+
+The objective function $Q(\theta)$ in the general case can have multiple local minima, and evaluating the value of $Q(.)$ at any given $\theta$ requires inversion of the matrix $\mathbf{K}_\theta + \sigma^{2}_n \mathbf{I}$ which has a time complexity $O(n^3)$ as noted above.
+In the interest of saving computational cost, one cannot use exhaustive search through the domain of the hyper-parameters to inform our choice for $\theta$.
+Some of the techniques used for model selection in the context of GPR include.
+
+\begin{enumerate}
+\item Grid Search: Construct a grid of values for $\theta$ as the cartesian product of one dimensional grids for each $\theta_i$, evaluate $Q(.)$ at each such grid point and choose the configuration which yields minimum value of $Q(.)$.
+
+\item Coupled Simulated Annealing: Introduced in \citet{Xavier-De-Souza2010}, it follows the same procedure as \emph{grid search}, but after evaluation of the objective $Q(.)$ on the grid, each grid point is iteratively mutated in a random walk fashion.
+This mutation is accepted or rejected according to the new value of $Q(.)$ as well as its value on the other grid points.
+This procedure is iterated until some stop criterion is reached.
+
+\item Maximum Likelihood: This technique as outlined in \citet{Rasmussen:2005:GPM:1162254} is a form of \emph{gradient descent}.
+It involves starting with an initial guess for $\theta$ and iteratively improving it by calculating the gradient of $Q(.)$ with respect to $\theta$.
+Although this method seems intuitive, it introduces an extra computational cost of calculating the gradient of $Q(\theta)$ with respect to each $\theta_i$ in every iteration and applying this method can sometimes lead to overfitting of the GPR model to the training data.
+
+\end{enumerate}
 
@@
 $Dst$ gives a measure of ring currents, which are modulated by plasma sheet particle injections into the inner magnetosphere during sub-storms.
-Studies have shown that the substorm occurrence rate increases with solar wind velocity (high speed streams) \citep{Kissinger2011,Newell2016}.
-Prolonged southward interplanetary magnetic field (IMF) $z$-component ($B_z$) is needed for sub-storms to occur \citep{McPherron1986}.
-An increase in the solar wind electric field, $VB_z$, can increase the dawn-dusk electric field in the magnetotail, which in turn determines the amount of plasma sheet particle that move to the inner magnetosphere \citep{Friedel2001}.
+Studies have shown that the substorm occurrence rate increases with solar wind velocity (high speed streams) \citet{Kissinger2011,Newell2016}.
+Prolonged southward interplanetary magnetic field (IMF) $z$-component ($B_z$) is needed for sub-storms to occur \citet{McPherron1986}.
+An increase in the solar wind electric field, $VB_z$, can increase the dawn-dusk electric field in the magnetotail, which in turn determines the amount of plasma sheet particle that move to the inner magnetosphere \citet{Friedel2001}.
 Therefore, our exogenous parameters consist of solar wind velocity and IMF $B_z$.
@@
 
+\subsection{Choice of Mean Function}
+
+Mean functions in GPR models encode trends in the data, they are the baseline predictions the model falls back to in case the training and test data have little correlation as predicted by the kernel function.
+If there is no prior knowledge about the function to be approximated, \citet{Rasmussen:2005:GPM:1162254} state that it is perfectly reasonable to choose $m(\mathbf{x} = 0)$ as the mean function, as long as the target values are normalized.
+In the case of the $Dst$ time series, it is known that the so called \emph{persistence model} $\hat{D}st(t) = Dst(t-1)$ performs quite well in the context of OSA prediction.
+We therefore choose the \emph{persistence model} as the mean function in our OSA Dst models.
+
 \subsection{Choice of Kernel}
 
-In this study, we construct Gaussian Process regression models with a combination of the \emph{maximum likelihood perceptron} kernel and \emph{student's T} kernel as shown in equations (\ref{eq:usedKernel}), for \emph{one step ahead} prediction of the $Dst$ index.
+In this study, we construct Gaussian Process regression models with a combination of the \emph{maximum likelihood perceptron} kernel and \emph{student's T} kernel as shown in equations (\ref{eq:usedKernel}).
+The \emph{maximum likelihood perceptron} kernel is the \emph{Gaussian Process} equivalent of a single hidden layer feed-forward neural network model as demonstrated in \citet{Neal:1996:BLN:525544}.
 

### 2017-02-15 agutmpl.tex
Added tables for validation and test sets
Resizing table font
Experiments section complete

@@
 \usepackage{graphicx}
+\usepackage{booktabs}
 \usepackage{url}
@@
 \usepackage{rotating}
+\usepackage{adjustbox}
 \graphicspath{{figures/}}
@@
 
-\section{Model Training and Validation} \label{sec:modeltraining}
+\section{Experiments} \label{sec:modeltraining}
 
-Before running performance bench marks for \emph{OSA} $Dst$ prediction on the storm events in \citet{Ji2012}, training and model selection of \emph{GP-AR} and \emph{GP-ARX} models on independent data sets must be performed.
-For this purpose we choose segments 00:00 1 January 2008 - 10:00 11 January 2008 for training and 00:00 15 November 2014 - 23:00 1 December 2014 for model selection.
+\subsection*{Training}
+
+We selected OMNI data sections 00:00 January 3 2010 to 23:00 January 23 2010 and 20:00 August 5 2011 to 22:00 August 6 2011 for training the GP-AR and GP-ARX models.
+The first training data section consists of ambient fluctuations of $Dst$ while the second contains a geomagnetic storm.
 
@@
 Note that this computation overhead is paid for every unique assignment to the model hyper-parameters.
-However, our chosen training set has a size of 250 which is still very much below the computational limits of the method and in our case.
+However, our chosen training set has a size of 243 which is still very much below the computational limits of the method and in our case.
 Solving equation \ref{eq:posteriormean} on a laptop computer takes less than a second for the training set considered in our analysis.
 
-\section{Experiments}\label{sec:exp}
+\subsection*{Selection}
+
+In order to find appropriate values of the hyper-parameters of the chosen kernel $K_osa(.,.)$, we apply \emph{grid search}, \emph{coupled simulated annealing} and \emph{maximum likelihood} methods.
+We fix the parameters $d$ and $\sigma^2$ of $K_st$ and model noise to values $0.01$ and $0.2$ respectively, the remaining parameters $w$ and $b$ are kept free to be calculated by model selection.
+Table \ref{table:modelselection} summarizes the settings used to run each model selection procedure.
+
+\subsection*{Validation}
+
+Apart from selecting the kernel parameters, one also needs to choose appropriate values for the auto-regressive orders $p$ in the case of GP-AR and $p, p_v, p_b$ in the case of GP-ARX.
+For this purpose we use a set of 24 storm events listed in table \ref{table:validationstorms} and for every assignment of values to the model order, we perform model selection with the routines in table \ref{table:modelselection} and record the performance on this validation set.
+
+In the case of GP-AR we let the model order $p$ vary from $5$ to $12$ while for GP-ARX we vary the total model order $p_t = p + p_v + p_b$ vary from $3$ to $12$ and for each $p_t$ evaluate every possible combination of $p$, $p_v$ and $p_b$ such that $p_t = p + p_v + p_b$ and $p, p_{v}, p_b > 0$.
+
+\subsection*{Evaluation}
+
+After selecting the best performing GP-AR and GP-ARX models in the validation phase, we test and compare the performance of these models with the predictions generated from the \emph{persistence model} $\hat{D}st(t) = Dst(t-1)$, on a set of 63 storm events occurring between 1998 and 2006 as given in table \ref{table:teststorms}.
 
@@
 
+\begin{table}[]
+\centering
+\caption{Settings of model selection procedures}
+\begin{tabular}{l c c c}
+\hline
+Procedure & Grid Size & Step & Max Iterations \\
+\hline
+Grid Search & 10 & 0.2 & NA \\
+Coupled Simulated Annealing & 4 & 0.2 & 30 \\
+Maximum likelihood & NA & 0.2 & 150\\
+\end{tabular}
+\label{table:modelselection}
+\end{table}
+
+\begin{table}[]
+\centering
+\caption{Storm events used for model selection of GP-AR and GP-ARX}
+\label{table:validationstorms}
+\begin{tabular}{llllll}
+\hline
+Event Id & Start Date & Start Hour & End Date & End Hour & min. Dst \\ \hline
+1 & 1995/03/26 & 0500 & 1995/03/26 & 2300 & −107 \\
+2 & 1995/04/07 & 1300 & 1995/04/09 & 0900 & −149 \\
+3 & 1995/09/27 & 0100 & 1995/09/28 & 0400 & −108 \\
+4 & 1995/10/18 & 1300 & 1995/10/19 & 1400 & −127 \\
+5 & 1996/10/22 & 2200 & 1996/10/23 & 1100 & −105 \\
+6 & 1997/04/21 & 1000 & 1997/04/22 & 0900 & −107 \\
+7 & 1997/05/15 & 0300 & 1997/05/16 & 0000 & −115 \\
+8 & 1997/10/10 & 1800 & 1997/10/11 & 1900 & −130 \\
+9 & 1997/11/07 & 0000 & 1997/11/07 & 1800 & −110 \\
+10 & 1997/11/22 & 2100 & 1997/11/24 & 0400 & −108 \\
+11 & 2005/06/12 & 1700 & 2005/06/13 & 1900 & −106 \\
+12 & 2005/08/31 & 1200 & 2005/09/01 & 1200 & −122 \\
+13 & 2006/12/14 & 2100 & 2006/12/16 & 0300 & −162 \\
+14 & 2011/09/26 & 1400 & 2011/09/27 & 1200 & −101 \\
+15 & 2011/10/24 & 2000 & 2011/10/25 & 1400 & −132 \\
+16 & 2012/03/08 & 1200 & 2012/03/10 & 1600 & −131 \\
+17 & 2012/04/23 & 1100 & 2012/04/24 & 1300 & −108 \\
+18 & 2012/07/15 & 0100 & 2012/07/16 & 2300 & −127 \\
+19 & 2012/09/30 & 1300 & 2012/10/01 & 1800 & −119 \\
+20 & 2012/10/08 & 0200 & 2012/10/09 & 1700 & −105 \\
+21 & 2012/11/13 & 1800 & 2012/11/14 & 1800 & −108 \\
+22 & 2013/03/17 & 0700 & 2013/03/18 & 1000 & −132 \\
+23 & 2013/05/31 & 1800 & 2013/06/01 & 2000 & −119 \\
+24 & 2014/02/18 & 1500 & 2014/02/19 & 1600 & −112
+\end{tabular}
+\end{table}
+
+\begin{table}[]
+\fontsize{9}{10.4}\selectfont
+\centering
+\caption{Storm events used to evaluate GP-AR and GP-ARX models}
+\label{table:teststorms}
+\begin{tabular}{cccccc}
+\hline
+Event Id & Start Date & Start Time & End Date & End Time & min. Dst \\ \hline
+1 & 1998/02/17 & 1200 & 1998/02/18 & 1000 & -100 \\
+2 & 1998/03/10 & 1100 & 1998/03/11 & 1800 & -116 \\
+3 & 1998/05/04 & 0200 & 1998/05/05 & 0200 & -205 \\
+4 & 1998/08/26 & 1000 & 1998/08/29 & 0700 & -155 \\
+5 & 1998/09/25 & 0100 & 1998/09/26 & 0000 & -207 \\
+6 & 1998/10/19 & 0500 & 1998/10/20 & 0800 & -112 \\
+7 & 1998/11/09 & 0300 & 1998/11/10 & 1600 & -142 \\
+8 & 1998/11/13 & 0000 & 1998/11/15 & 0400 & -131 \\
+9 & 1999/01/13 & 1600 & 1999/01/14 & 2000 & -112 \\
+10 & 1999/02/18 & 0300 & 1999/02/19 & 2100 & -123 \\
+11 & 1999/09/22 & 2000 & 1999/09/23 & 2300 & -173 \\
+12 & 1999/10/22 & 0000 & 1999/10/23 & 1400 & -237 \\
+13 & 2000/02/12 & 0500 & 2000/02/13 & 1500 & -133 \\
+14 & 2000/04/06 & 1700 & 2000/04/08 & 0900 & -288 \\
+15 & 2000/05/24 & 0100 & 2000/05/25 & 2000 & -147 \\
+16 & 2000/08/10 & 2000 & 2000/08/11 & 1800 & -106 \\
+17 & 2000/08/12 & 0200 & 2000/08/13 & 1700 & -235 \\
+18 & 2000/10/13 & 0200 & 2000/10/14 & 2300 & -107 \\
+19 & 2000/10/28 & 2000 & 2000/10/29 & 2000 & -127 \\
+20 & 2000/11/06 & 1300 & 2000/11/07 & 1800 & -159 \\
+21 & 2000/11/28 & 1800 & 2000/11/29 & 2300 & -119 \\
+22 & 2001/03/19 & 1500 & 2001/03/21 & 2300 & -149 \\
+23 & 2001/03/31 & 0400 & 2001/04/01 & 2100 & -387 \\
+24 & 2001/04/11 & 1600 & 2001/04/13 & 0700 & -271 \\
+25 & 2001/04/18 & 0100 & 2001/04/18 & 1300 & -114 \\
+26 & 2001/04/22 & 0200 & 2001/04/23 & 1500 & -102 \\
+27 & 2001/08/17 & 1600 & 2001/08/18 & 1600 & -105 \\
+28 & 2001/09/30 & 2300 & 2001/10/02 & 0000 & -148 \\
+29 & 2001/10/21 & 1700 & 2001/10/24 & 1100 & -187 \\
+30 & 2001/10/28 & 0300 & 2001/10/29 & 2200 & -157 \\
+31 & 2002/03/23 & 1400 & 2002/03/25 & 0500 & -100 \\
+32 & 2002/04/17 & 1100 & 2002/04/19 & 0200 & -127 \\
+33 & 2002/04/19 & 0900 & 2002/04/21 & 0600 & -149 \\
+34 & 2002/05/11 & 1000 & 2002/05/12 & 1600 & -110 \\
+35 & 2002/05/23 & 1200 & 2002/05/24 & 2300 & -109 \\
+36 & 2002/08/01 & 2300 & 2002/08/02 & 0900 & -102 \\
+37 & 2002/09/04 & 0100 & 2002/09/05 & 0000 & -109 \\
+38 & 2002/09/07 & 1400 & 2002/09/08 & 2000 & -181 \\
+39 & 2002/10/01 & 0600 & 2002/10/03 & 0800 & -176 \\
+40 & 2002/10/03 & 1000 & 2002/10/04 & 1800 & -146 \\
+41 & 2002/11/20 & 1600 & 2002/11/22 & 0600 & -128 \\
+42 & 2003/05/29 & 2000 & 2003/05/30 & 1000 & -144 \\
+43 & 2003/06/17 & 1900 & 2003/06/19 & 0300 & -141 \\
+44 & 2003/07/11 & 1500 & 2003/07/12 & 1600 & -105 \\
+45 & 2003/08/17 & 1800 & 2003/08/19 & 1100 & -148 \\
+46 & 2003/11/20 & 1200 & 2003/11/22 & 0000 & -422 \\
+47 & 2004/01/22 & 0300 & 2004/01/24 & 0000 & -149 \\
+48 & 2004/02/11 & 1000 & 2004/02/12 & 0000 & -105 \\
+49 & 2004/04/03 & 1400 & 2004/04/04 & 0800 & -112 \\
+50 & 2004/07/22 & 2000 & 2004/07/23 & 2000 & -101 \\
+51 & 2004/07/24 & 2100 & 2004/07/26 & 1700 & -148 \\
+52 & 2004/07/26 & 2200 & 2004/07/30 & 0500 & -197 \\
+53 & 2004/08/30 & 0500 & 2004/08/31 & 2100 & -126 \\
+54 & 2004/11/07 & 2100 & 2004/11/08 & 2100 & -373 \\
+55 & 2004/11/09 & 1100 & 2004/11/11 & 0900 & -289 \\
+56 & 2004/11/11 & 2200 & 2004/11/13 & 1300 & -109 \\
+57 & 2005/01/21 & 1800 & 2005/01/23 & 0500 & -105 \\
+58 & 2005/05/07 & 2000 & 2005/05/09 & 1000 & -127 \\
+59 & 2005/05/29 & 2200 & 2005/05/31 & 0800 & -138 \\
+60 & 2005/06/12 & 1700 & 2005/06/13 & 1900 & -106 \\
+61 & 2005/08/31 & 1200 & 2005/09/01 & 1200 & -131 \\
+62 & 2006/04/13 & 2000 & 2006/04/14 & 2300 & -111 \\
+63 & 2006/12/14 & 2100 & 2006/12/16 & 0300 & -147 \\ \hline
+\end{tabular}%
+\end{table}
+
 % & $b \in \mathbb{R}, d \in \mathbb{N}$  & $l \in \mathbb{R}$

### 2017-02-15 agutmpl.tex
Added table for evaluation results
Manuscript iteration 0

@@
 
+For measuring performance of model instances on the validation set storm events, the \emph{mean absolute error} (MAE), \emph{root mean square error} (RMSE) and \emph{coefficient of correlation} (CC) are calculated.
+
 In the case of GP-AR we let the model order $p$ vary from $5$ to $12$ while for GP-ARX we vary the total model order $p_t = p + p_v + p_b$ vary from $3$ to $12$ and for each $p_t$ evaluate every possible combination of $p$, $p_v$ and $p_b$ such that $p_t = p + p_v + p_b$ and $p, p_{v}, p_b > 0$.
@@
 
-\section{Conclusions}\label{sec:res}
+\section{Results}\label{sec:res}
+
+Figures \ref{fig:CompareMae} and \ref{fig:CompareCC} show how the mean absolute error and coefficient of correlation as calculated on the validation set of table \ref{table:validationstorms}, vary with increasing model order for GP-AR and GP-ARX.
+In both cases the results first improve and then stagnate or worsen.
+
+Figures \ref{fig:CompareMaeARX} and \ref{fig:CompareCCARX} break down the results for GP-ARX by the model selection routine used.
+Apart from the general trend observed in \ref{fig:CompareMAE} and \ref{fig:CompareCC}, we also observe that \emph{grid search} and \emph{coupled simulated annealing} give superior performance as compared to gradient based \emph{maximum likelihood}.
+
+From the validation results, we can chose the model order which yields the best performance, for GP-AR it is $p_t = 6$ while for GP-ARX it is $p_t = 11$.
+Further examination of the validation results shows that in the scheme $p_t = 11$ choosing $p = 7, p_v = 1, p_b = 3$ gives superior results.
+
+After choosing the best performing GP-AR and GP-ARX models, we calculate their performance on the test set of table \ref{table:teststorms}.
+The results of these model evaluations are summarized in table \ref{table:results}, the GP-AR and GP-ARX models improve upon the performance of the \emph{persistence model}.
+
+Figures \ref{fig:ComparePred1}, \ref{fig:ComparePred2} and \ref{fig:ComparePred3} show OSA predictions of the GP-ARX model with $\pm \sigma$ error bars for three storm events in the years 1998 and 2000.
+The GP-ARX model gives accurate predictions along with plausible error bars around its mean predictions.
+
+\section{Conclusions}
+
+In this paper, we describe a flexible and expressive methodology for generating probabilistic forecasts of the $Dst$ index.
+We proposed two \emph{Gaussian Process} auto-regressive models, \emph{GP-ARX} and \emph{GP-AR}, to generate hourly predictions and their associated error bars.
+We also describe how to carry out model selection and validation of GP-AR and GP-ARX models.
+
+Our results can be summarized as follows.
+\begin{enumerate}
+      \item \emph{Persistence} model in the model building and evaluation process in the context of \emph{one step ahead} prediction of the $Dst$ index.
+It is clear that the \emph{persistence} behavior in the $Dst$ values is very strong i.e.
+the trivial predictive model $\hat{Dst}(t) = Dst(t-1)$ gives excellent performance according to the metrics chosen.
+
+      \item \emph{Gaussian Process} AR and ARX models give encouraging benefits in OSA prediction.
+Leveraging the strengths of the Bayesian approach, they are able to learn robust predictors from data.
+If one compares the training sets used by all the models, one can appreciate that the models presented here need relatively small training and validations sets: the training set contains 243 instances, while the validation set contains 782 instances.
+
+      \item Since the GP models generate predictive distributions for test data and not just point predictions they lend themselves to the requirements of space weather prediction very well because of the need to generate error bars on predictions.
+\end{enumerate}
 
@@
 \noindent\includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm5.png}
-\caption{OSA Predictions with $\pm \sigma$ error bars for event: }
+\caption{OSA Predictions with $\pm \sigma$ error bars for event: 1998/09/25 01:00 to 1998/09/26 00:00}
 \label{fig:ComparePred1}
@@
 \noindent\includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm8.png}
-\caption{OSA Predictions with $\pm \sigma$ error bars for event: }
+\caption{OSA Predictions with $\pm \sigma$ error bars for event: 1998/11/13 00:00 to 1998/11/15 04:00}
 \label{fig:ComparePred2}
@@
 \noindent\includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm14.png}
-\caption{OSA Predictions with $\pm \sigma$ error bars for event: }
-\label{fig:ComparePred54}
+\caption{OSA Predictions with $\pm \sigma$ error bars for event: 2000/04/06 17:00 to 2000/04/08 09:00}
+\label{fig:ComparePred3}
 \end{figure}
@@
 \label{table:modelselection}
+\end{table}
+
+\begin{table}[]
+\centering
+\caption{Evaluation results for models on storm events listed in table \ref{table:teststorms}}
+\label{table:results}
+\begin{tabular}{l c c c}
+\hline
+Model & Mean Absolute Error & Root Mean Square Error & Coefficient of Correlation\\ \hline
+GP-ARX & 7.252 & 11.93 & 0.972\\
+GP-AR & 8.37 & 14.04 & 0.963\\
+Persistence & 9.182 & 14.94 & 0.957\\
+\end{tabular}
 \end{table}

### 2017-03-01 agutmpl.tex
Updates from ShareLaTeX

@@
 %   when using Draft:
-%  \setkeys{Gin}{draft=false}
+  \setkeys{Gin}{draft=false}
 %
@@
 %Talk about need for probabilistic forecasts.
-Although much research has been done on prediction of the $Dst$ index, must less has been done on probabilistic forecasting of $Dst$.
-One such work described in \citet{McPherron:2013} involves identification of high speed solar wind streams using the W-S-A model (see \citet{WSAModel}), using predictions of high speed streams to construct ensembles of $Dst$ trajectories which yield the quartiles of $Dst$ time series.
+Although much research has been done on prediction of the $Dst$ index, much less has been done on probabilistic forecasting of $Dst$.
+One such work described in \citet{McPherron:2013} involves identification of high speed solar wind streams using the WSA model (see \citet{WSAModel}), using predictions of high speed streams to construct ensembles of $Dst$ trajectories which yield the quartiles of $Dst$ time series.
 
-In this work we propose an technique for probabilistic forecating of $Dst$, which yields a predictive distribution as a closed form expression.
-Our models take as input past values of $Dst$, solar wind speed and the z-component of the \emph{Interplanetary Magnetic Field} (IMF) and output a Gaussian distribution with a specific mean and variance as the OSA prediction of the $Dst$.
+In this work we propose a technique for probabilistic forecasting of $Dst$, which yields a predictive distribution as a closed form expression.
+Our models take as input past values of $Dst$, solar wind speed and the \textit{z} component of the \emph{Interplanetary Magnetic Field} (IMF) and output a Gaussian distribution with a specific mean and variance as the OSA prediction of the $Dst$.
 
@@
 The quantities $l$ in the RBF, and $b$ and $d$ in the polynomial kernel are known as \emph{hyper-parameters}.
-Hyper-parameters give flexibility to particular kernel structure, for example $d = 1, 2, 3, \cdots$ in the polynomial kernel represents linear, quadratic, cubic and higher order polynomials respectively.
+Hyper-parameters give flexibility to a particular kernel structure, for example $d = 1, 2, 3, \cdots$ in the polynomial kernel represents linear, quadratic, cubic and higher order polynomials respectively.
 The process of assigning values to the \emph{hyper-parameters} is crucial in the model building process and is known as \emph{model selection}.
@@
 Note that this computation overhead is paid for every unique assignment to the model hyper-parameters.
-However, our chosen training set has a size of 243 which is still very much below the computational limits of the method and in our case.
-Solving equation \ref{eq:posteriormean} on a laptop computer takes less than a second for the training set considered in our analysis.
+However, our chosen training set has a size of 243 which is still very much below the computational limits of the method and in our case solving equation \ref{eq:posteriormean} on a laptop computer takes less than a second for the training set considered in our analysis.
 
@@
 
-In order to find appropriate values of the hyper-parameters of the chosen kernel $K_osa(.,.)$, we apply \emph{grid search}, \emph{coupled simulated annealing} and \emph{maximum likelihood} methods.
-We fix the parameters $d$ and $\sigma^2$ of $K_st$ and model noise to values $0.01$ and $0.2$ respectively, the remaining parameters $w$ and $b$ are kept free to be calculated by model selection.
+In order to find appropriate values of the hyper-parameters of the chosen kernel $K_{osa}$, we apply \emph{grid search}, \emph{coupled simulated annealing} and \emph{maximum likelihood} methods.
+We fix the parameters $d$ and $\sigma^2$ of $K_{st}$ and model noise to values $0.01$ and $0.2$ respectively, the remaining parameters $w$ and $b$ are kept free to be calculated by model selection.
 Table \ref{table:modelselection} summarizes the settings used to run each model selection procedure.
@@
 
-For measuring performance of model instances on the validation set storm events, the \emph{mean absolute error} (MAE), \emph{root mean square error} (RMSE) and \emph{coefficient of correlation} (CC) are calculated.
+For measuring performance of model instances on the validation set storm events, the following metrics are calculated.
+
+\begin{enumerate}
+    \item The mean absolute error.
+    \begin{equation}
+        MAE = \sum_{t=1}^{n} \left |(Dst(t) - \hat{D}st(t)) \right | / n
+    \end{equation}
+    \item The root mean square error.
+    \begin{equation}
+        RMSE = \sqrt{\sum_{t=1}^{n} (Dst(t) - \hat{D}st(t))^2 / n}
+    \end{equation}
+    \item Correlation coefficient between the predicted and actual value of $Dst$.
+    \begin{equation}
+        CC = Cov(Dst, \hat{D}st)/\sqrt{Var(Dst) Var(\hat{D}st)}
+    \end{equation}
+\end{enumerate}
 
@@
 Figures \ref{fig:CompareMaeARX} and \ref{fig:CompareCCARX} break down the results for GP-ARX by the model selection routine used.
-Apart from the general trend observed in \ref{fig:CompareMAE} and \ref{fig:CompareCC}, we also observe that \emph{grid search} and \emph{coupled simulated annealing} give superior performance as compared to gradient based \emph{maximum likelihood}.
+Apart from the general trend observed in \ref{fig:CompareMae} and \ref{fig:CompareCC}, we also observe that \emph{grid search} and \emph{coupled simulated annealing} give superior performance as compared to gradient based \emph{maximum likelihood}.
 

### 2017-03-02 agutmpl.tex
Modifications Iteration: 1

@@
 
-After selecting the best performing GP-AR and GP-ARX models in the validation phase, we test and compare the performance of these models with the predictions generated from the \emph{persistence model} $\hat{D}st(t) = Dst(t-1)$, on a set of 63 storm events occurring between 1998 and 2006 as given in table \ref{table:teststorms}.
+After selecting the best performing GP-AR and GP-ARX models in the validation phase, we test and compare the performance of these models with the predictions generated from the \emph{persistence model} $\hat{D}st(t) = Dst(t-1)$, on a set of 63 storm events occurring between 1998 and 2006 as given in table \ref{table:teststorms}, which is the same list of storm events as used in \citet{Ji2012}.
 
@@
 
-Figures \ref{fig:CompareMae} and \ref{fig:CompareCC} show how the mean absolute error and coefficient of correlation as calculated on the validation set of table \ref{table:validationstorms}, vary with increasing model order for GP-AR and GP-ARX.
-In both cases the results first improve and then stagnate or worsen.
+Figures \ref{fig:CompareMae} and \ref{fig:CompareCC} show how the mean absolute error and coefficient of correlation as calculated on the validation set storm events of table \ref{table:validationstorms}, vary with increasing model order for GP-AR and GP-ARX.
+The results are represented as boxplots, in which a rectangle is drawn to represent the second and third quartiles, with a vertical line inside to indicate the median value while outlying points are shown as dots.
+In both cases, the predictive performance first improves and then stagnates or worsens with increasing model order.
 
@@
 
-Figures \ref{fig:ComparePred1}, \ref{fig:ComparePred2} and \ref{fig:ComparePred3} show OSA predictions of the GP-ARX model with $\pm \sigma$ error bars for three storm events in the years 1998 and 2000.
+Figures \ref{fig:ComparePred1}, \ref{fig:ComparePred2} and \ref{fig:ComparePred3} show OSA predictions of the GP-ARX model with $\pm \sigma$ error bars for three storm events in the time period between 1998 and 2003.
 The GP-ARX model gives accurate predictions along with plausible error bars around its mean predictions.
@@
 \begin{enumerate}
-      \item \emph{Persistence} model in the model building and evaluation process in the context of \emph{one step ahead} prediction of the $Dst$ index.
-It is clear that the \emph{persistence} behavior in the $Dst$ values is very strong i.e.
+      \item \emph{Persistence} model plays an important role in the model building and evaluation process in the context of \emph{one step ahead} prediction of the $Dst$ index.
+It is clear that the persistence behavior in the $Dst$ values is very strong i.e.
 the trivial predictive model $\hat{Dst}(t) = Dst(t-1)$ gives excellent performance according to the metrics chosen.
@@
       \item Since the GP models generate predictive distributions for test data and not just point predictions they lend themselves to the requirements of space weather prediction very well because of the need to generate error bars on predictions.
+
+      \item The \emph{Gaussian Process} regression framework described in this study can also be extended to multiple hour ahead prediction of $Dst$, which is currently a work in progress.
 \end{enumerate}
@@
 \noindent\includegraphics[width=\textwidth]{Compare-mae.png}
-\caption{Mean Absolute Error on validation set storms vs model order for GP-AR and GP-ARX}
+\caption{Mean Absolute Error on validation set storms vs model order for GP-AR and GP-ARX.
+\\ \textbf{Key}: Rectangle borders represent the second and third quartiles, with a horizontal line inside to indicate the median value while outlying points are shown as dots}
 \label{fig:CompareMae}
@@
 \noindent\includegraphics[width=\textwidth]{Compare-cc.png}
-\caption{Mean Absolute Error on validation set storms vs model order for GP-AR and GP-ARX}
+\caption{Coefficient of Correlation on validation set storms vs model order for GP-AR and GP-ARX \\ \textbf{Key}: Rectangle borders represent the second and third quartiles, with a horizontal line inside to indicate the median value while outlying points are shown as dots}
 \label{fig:CompareCC}
@@
 \noindent\includegraphics[width=\textwidth]{Compare-mae-arx.png}
-\caption{Mean Absolute Error on validation set storms vs model order for GP-AR and GP-ARX}
+\caption{Mean Absolute Error on validation set storms vs model order for GP-AR and GP-ARX for \emph{CSA}, \emph{GS} and \emph{ML} model selection routines \\ Rectangle borders represent the second and third quartiles, with a horizontal line inside to indicate the median value while outlying points are shown as dots}
 \label{fig:CompareMaeARX}
@@
 \noindent\includegraphics[width=\textwidth]{Compare-cc-arx.png}
-\caption{Mean Absolute Error on validation set storms vs model order for GP-AR and GP-ARX}
+\caption{Coefficient of Correlation on validation set storms vs model order for GP-AR and GP-ARX for \emph{CSA}, \emph{GS} and \emph{ML} model selection routines \\ Rectangle borders represent the second and third quartiles, with a horizontal line inside to indicate the median value while outlying points are shown as dots}
 \label{fig:CompareCCARX}
@@
 \begin{figure}
-\noindent\includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm5.png}
-\caption{OSA Predictions with $\pm \sigma$ error bars for event: 1998/09/25 01:00 to 1998/09/26 00:00}
+\noindent\includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm43.png}
+\caption{OSA Predictions with $\pm \sigma$ error bars for event: 2003/06/17 19:00 to 2003/06/19 03:00}
 \label{fig:ComparePred1}
@@
 \begin{figure}
-\noindent\includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm14.png}
-\caption{OSA Predictions with $\pm \sigma$ error bars for event: 2000/04/06 17:00 to 2000/04/08 09:00}
+\noindent\includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm9.png}
+\caption{OSA Predictions with $\pm \sigma$ error bars for event: 1999/01/13 16:00 to 1999/01/14 20:00}
 \label{fig:ComparePred3}

### 2017-03-11 agutmpl.tex
Updates from ShareLaTeX

@@
 Geomagnetic indices come in various forms, they may take continuous or discrete values and may be defined with varying time resolutions.
-Their values are often calculated by averaging or combining a number of readings taken by instruments around the Earth.
+Their values are often calculated by averaging or combining a number of readings taken by instruments, usually magnetometers, around the Earth.
 Each geomagnetic index is a proxy for a particular kind of phenomenon.
@@
 \begin{enumerate}
-    \item $K_p$: The Kp-index is a discrete valued global geomagnetic storm index and is based on 3 hour measurements of the K-indices \citet{Bartels}.
+    \item $K_p$: The Kp-index is a discrete valued global geomagnetic activity index and is based on 3 hour measurements of the K-indices \citet{Bartels}.
 The K-index itself is a three hour long quasi-logarithmic local index of the geomagnetic activity, relative to a calm day curve for the given location.
@@
 
-    \item $Dst$: A continuous hourly index which measures the weakening of the Earths magnetic field due to ring currents and the strength of geomagnetic storms \citet{DesslerAndParker}. 
+    \item $Dst$: A continuous hourly index which gives a measure of the weakening or streghtening of the Earth's equatorial magnetic field due to the weakening or strenghttening of the ring currents and the geomagnetic storms \citet{DesslerAndParker}.
 \end{enumerate}
@@
 The NARMAX methodology builds models by constructing polynomial expansions of inputs and determines the best combinations of monomials to include in the refined model by using a criterion called the \emph{error reduction ratio} (ERR).
-The parameters of the so called NARMAX OLS-ERR model are calculated by solving the \emph{ordinary least squares} (OLS) problem arising from a quadratic objective objective function.
+The parameters of the so called NARMAX OLS-ERR model are calculated by solving the \emph{ordinary least squares} (OLS) problem arising from a quadratic objective function.
 The reader may refer to \citet{billings2013nonlinear} for a detailed exposition of the NARMAX methodology.
@@
 
-Let us assume that we want to model a process in which a scalar quantity $y$ is described by $y = f(\mathbf{x}) + \epsilon$ where   $f(.): \mathbb{R}^d \rightarrow \mathbb{R}$ is an unknown scalar function of a multidimensional input vector $\mathbf{x} \in \mathbb{R}^d$, and $\epsilon \sim \mathcal{N}(0, \sigma^2)$ is Gaussian distributed noise with variance $\sigma^2$.
+Let us assume that we want to model a process in which a scalar quantity $y$ is specified as $y = f(\mathbf{x}) + \epsilon$ where   $f(.): \mathbb{R}^d \rightarrow \mathbb{R}$ is an unknown scalar function of a multidimensional input vector $\mathbf{x} \in \mathbb{R}^d$, $d$ is the dimensionality of the input space, and $\epsilon \sim \mathcal{N}(0, \sigma^2)$ is Gaussian distributed noise with variance $\sigma^2$.
 
@@
 
-\begin{table}
+\begin{table}[h]
 \caption{Popular Kernel functions used in GPR models}
@@
 
-\begin{table}[]
+\begin{table}[h]
 \centering
@@
 
-\begin{table}[]
+\begin{table}[h]
 \fontsize{9}{10.4}\selectfont

### 2017-03-16 agutmpl.tex
Updates from ShareLaTeX

@@
 \begin{enumerate}
-    \item $K_p$: The Kp-index is a discrete valued global geomagnetic activity index and is based on 3 hour measurements of the K-indices \citet{Bartels}.
+    \item $K_p$: The Kp-index is a discrete valued global geomagnetic activity index and is based on 3 hour measurements of the K-indices \citep{Bartels}.
 The K-index itself is a three hour long quasi-logarithmic local index of the geomagnetic activity, relative to a calm day curve for the given location.
 
-    \item $AE$: The Auroral Electrojet Index, $AE$, is designed to provide a global, quantitative measure of auroral zone magnetic activity produced by enhanced Ionospheric currents flowing below and within the auroral oval \citet{AEIndex}.
+    \item $AE$: The Auroral Electrojet Index, $AE$, is designed to provide a global, quantitative measure of auroral zone magnetic activity produced by enhanced Ionospheric currents flowing below and within the auroral oval \citep{AEIndex}.
 It is a continuous index which is calculated every hour.
 
-    \item $Dst$: A continuous hourly index which gives a measure of the weakening or streghtening of the Earth's equatorial magnetic field due to the weakening or strenghttening of the ring currents and the geomagnetic storms \citet{DesslerAndParker}.
+    \item $Dst$: A continuous hourly index which gives a measure of the weakening or strengthening of the Earth's equatorial magnetic field due to the weakening or strengthening of the ring currents and the geomagnetic storms \citep{DesslerAndParker}.
 \end{enumerate}

### 2017-03-16 agutmpl.tex
Penultimate changes

@@
 For the present study, we focus on prediction of the hourly $Dst$ index which is a straightforward indicator of geomagnetic storms.
-More specifically, we focus on the \emph{one step ahead} (OSA), in this case one hour ahead prediction of $Dst$ because it is the simplest model towards building long term predictions of geomagnetic response of the Earth to changing space weather conditions.
+More specifically, we focus on the \emph{one step ahead} (OSA) (in this case one hour ahead) prediction of $Dst$ because it is the simplest model towards building long term predictions of geomagnetic response of the Earth to changing space weather conditions.
 
 The $Dst$ OSA prediction problem has been the subject of several modeling efforts in the literature.
-One of the earliest models has been presented by \citet{JGR:JGR10260} who calculated $Dst(t)$ as the solution of an \emph{Ordinary Differential Equation} (ODE) which expressed the rate of change of $Dst(t)$ as a combination of two terms: decay and injection $\frac{d Dst(t)}{dt} = Q(t) - \frac{Dst(t)}{\tau}$, where Q(t) relates to the particle injection from the plasma sheet into the inner magnetosphere.
+One of the earliest models has been presented by \citet{JGR:JGR10260} who calculated $Dst(t)$ as the solution of an \emph{Ordinary Differential Equation} (ODE) which expressed the rate of change of $Dst(t)$ as a combination of two terms: decay and injection $\frac{d Dst(t)}{dt} = Q(t) - \frac{Dst(t)}{\tau}$, where $Q(t)$ relates to the particle injection from the plasma sheet into the inner magnetosphere.
 
@@
 
-Let us assume that we want to model a process in which a scalar quantity $y$ is specified as $y = f(\mathbf{x}) + \epsilon$ where   $f(.): \mathbb{R}^d \rightarrow \mathbb{R}$ is an unknown scalar function of a multidimensional input vector $\mathbf{x} \in \mathbb{R}^d$, $d$ is the dimensionality of the input space, and $\epsilon \sim \mathcal{N}(0, \sigma^2)$ is Gaussian distributed noise with variance $\sigma^2$.
+Let us assume that we want to model a process in which a scalar quantity $y$ is specified as $y = f(\mathbf{x}) + \epsilon$ where   $f(.): \mathbb{R}^d \rightarrow \mathbb{R}$ is an unknown scalar function of a multidimensional input vector $\mathbf{x} \in \mathbb{R}^d$, $d$ is the dimensionality of the input space, and $\epsilon \sim \mathcal{N}(0, \sigma^2)$ is zero mean Gaussian noise with variance $\sigma^2$.
 
@@
 \begin{align}
-  \mathbf{X}  = & \left( \begin{array}{c} \mathbf{x}^{T}_1 \\ \mathbf{x}^{T}_2 \\ \vdots \\ \mathbf{x}^{T}_n \end{array} \right)_{n \times d} \label{eq:feat} \\
+  \mathbf{X}  = & \left( \begin{array}{c} \mathbf{x}^{T}_1 \\ \mathbf{x}^{T}_2 \\ \vdots \\ \mathbf{x}^{T}_N \end{array} \right)_{N \times d} \label{eq:feat} \\
   \vspace{2\baselineskip}
-  \mathbf{y}  = & \left( \begin{array}{c} y_1 \\ y_2 \\ \vdots \\ y_N \end{array} \right) _{n \times 1} \label{eq:labels}
+  \mathbf{y}  = & \left( \begin{array}{c} y_1 \\ y_2 \\ \vdots \\ y_N \end{array} \right) _{N \times 1} \label{eq:labels}
 \end{align}
@@
 \begin{enumerate}
-      \item $\mathbf{I}$: The $n \times n$ identity matrix.
-      \item $\mathbf{K} = [K(\mathbf{x}_i, \mathbf{x}_j)], \ i,j \in 1,\cdots,n$ : Kernel matrix constructed from all couples obtained from the training data.
-      \item $\mathbf{K}_{*} = [K(\mathbf{x}_i, \mathbf{x}^{*}_j)], \ i \in 1,\cdots,n ; j \in 1,\cdots,m$ : Cross kernel matrix constructed from all couples between training and test data points.
-      \item $\mathbf{K}_{**} = [K(\mathbf{x}^{*}_i, \mathbf{x}^{*}_j)], \ i,j \in 1,\cdots,m$: Kernel matrix constructed from all couples obtained from the test data.
+      \item $\mathbf{I}$: The $N \times N$ identity matrix.
+      \item $\mathbf{K} = [K(\mathbf{x}_i, \mathbf{x}_j)], \ i,j \in 1,\cdots,N$ : Kernel matrix constructed from all couples obtained from the training data.
+      \item $\mathbf{K}_{*} = [K(\mathbf{x}_i, \mathbf{x}^{*}_j)], \ i \in 1,\cdots,N ; j \in 1,\cdots,M$ : Cross kernel matrix constructed from all couples between training and test data points.
+      \item $\mathbf{K}_{**} = [K(\mathbf{x}^{*}_i, \mathbf{x}^{*}_j)], \ i,j \in 1,\cdots,M$: Kernel matrix constructed from all couples obtained from the test data.
 \end{enumerate}
@@
 The interested reader is suggested to refer to \citet{Berlinet2004}, \citet{Scholkopf:2001:LKS:559923} and \citet{hofmann2008} for a thorough treatment of kernel functions and the rich theory behind them.
-Some common kernel functions used in machine learning are listed in table \ref{table:kernel}.
+Some common kernel functions used in machine learning are listed in Table \ref{table:kernel}.
 
@@
   Q(\theta) & = - log(p(\mathbf{y}|\mathbf{X}, K_\theta)) \\
-            & = -\frac{1}{2} \mathbf{y}^\intercal (\mathbf{K}_\theta + \sigma^{2}_n \mathbf{I})^{-1} \mathbf{y} - \frac{1}{2}|\mathbf{K}_\theta + \sigma^{2}_n \mathbf{I}| - \frac{n}{2}log(2\pi) \\
-  \mathbf{K}_\theta & = [K_{\theta}(\mathbf{x}_i, \mathbf{x}_j)]_{n \times n}
+            & = -\frac{1}{2} \mathbf{y}^\intercal (\mathbf{K}_\theta + \sigma^{2} \mathbf{I})^{-1} \mathbf{y} - \frac{1}{2}|\mathbf{K}_\theta + \sigma^{2} \mathbf{I}| - \frac{N}{2}log(2\pi) \\
+  \mathbf{K}_\theta & = [K_{\theta}(\mathbf{x}_i, \mathbf{x}_j)]_{N \times N}
 \end{align*}
@@
 
-The objective function $Q(\theta)$ in the general case can have multiple local minima, and evaluating the value of $Q(.)$ at any given $\theta$ requires inversion of the matrix $\mathbf{K}_\theta + \sigma^{2}_n \mathbf{I}$ which has a time complexity $O(n^3)$ as noted above.
+The objective function $Q(\theta)$ in the general case can have multiple local minima, and evaluating the value of $Q(.)$ at any given $\theta$ requires inversion of the matrix $\mathbf{K}_\theta + \sigma^{2} \mathbf{I}$ which has a time complexity $O(N^3)$ as noted above.
 In the interest of saving computational cost, one cannot use exhaustive search through the domain of the hyper-parameters to inform our choice for $\theta$.
@@
 It involves starting with an initial guess for $\theta$ and iteratively improving it by calculating the gradient of $Q(.)$ with respect to $\theta$.
-Although this method seems intuitive, it introduces an extra computational cost of calculating the gradient of $Q(\theta)$ with respect to each $\theta_i$ in every iteration and applying this method can sometimes lead to overfitting of the GPR model to the training data.
+Although this method seems intuitive, it introduces an extra computational cost of calculating the gradient of $Q(\theta)$ with respect to each $\theta_i$ in every iteration and applying this method can sometimes lead to overfitting of the GPR model to the training data \citep{Rasmussen:2005:GPM:1162254}.
 
@@
 Prolonged southward interplanetary magnetic field (IMF) $z$-component ($B_z$) is needed for sub-storms to occur \citet{McPherron1986}.
-An increase in the solar wind electric field, $VB_z$, can increase the dawn-dusk electric field in the magnetotail, which in turn determines the amount of plasma sheet particle that move to the inner magnetosphere \citet{Friedel2001}.
-Therefore, our exogenous parameters consist of solar wind velocity and IMF $B_z$.
+An increase in the solar wind electric field, $V_{sw}B_z$, can increase the dawn-dusk electric field in the magnetotail, which in turn determines the amount of plasma sheet particle that move to the inner magnetosphere \citet{Friedel2001}.
+Therefore, our exogenous parameters consist of solar wind velocity $V_{sw}$ and IMF $B_z$.
 
@@
        \mathbf{x}_{t-1} & = (Dst(t-1), \cdots , Dst(t-p+1), \\
-        & \ \ \ \ \  V(t-1), \cdots, V(t-p_{v}+1),\\
+        & \ \ \ \ \  V_{sw}(t-1), \cdots, V_{sw}(t-p_{v}+1),\\
         & \ \ \ \ \  B_{z}(t-1), \cdots, B_{z}(t-p_{b}+1))
@@
 Apart from selecting the kernel parameters, one also needs to choose appropriate values for the auto-regressive orders $p$ in the case of GP-AR and $p, p_v, p_b$ in the case of GP-ARX.
-For this purpose we use a set of 24 storm events listed in table \ref{table:validationstorms} and for every assignment of values to the model order, we perform model selection with the routines in table \ref{table:modelselection} and record the performance on this validation set.
+For this purpose we use a set of 24 storm events listed in Table \ref{table:validationstorms} and for every assignment of values to the model order, we perform model selection with the routines in Table \ref{table:modelselection} and record the performance on this validation set.
 
@@
 
-In the case of GP-AR we let the model order $p$ vary from $5$ to $12$ while for GP-ARX we vary the total model order $p_t = p + p_v + p_b$ vary from $3$ to $12$ and for each $p_t$ evaluate every possible combination of $p$, $p_v$ and $p_b$ such that $p_t = p + p_v + p_b$ and $p, p_{v}, p_b > 0$.
+In the case of GP-AR we let the model order $p$ vary from $5$ to $12$ while for GP-ARX we let the total model order $p_t = p + p_v + p_b$ vary from $3$ to $12$ and for each $p_t$ evaluate every possible combination of $p$, $p_v$ and $p_b$ such that $p_t = p + p_v + p_b$ and $p, p_{v}, p_b > 0$.
 
@@
 
-After selecting the best performing GP-AR and GP-ARX models in the validation phase, we test and compare the performance of these models with the predictions generated from the \emph{persistence model} $\hat{D}st(t) = Dst(t-1)$, on a set of 63 storm events occurring between 1998 and 2006 as given in table \ref{table:teststorms}, which is the same list of storm events as used in \citet{Ji2012}.
+After selecting the best performing GP-AR and GP-ARX models in the validation phase, we test and compare the performance of these models with the predictions generated from the \emph{persistence model} $\hat{D}st(t) = Dst(t-1)$, on a set of 63 storm events occurring between 1998 and 2006 as given in Table \ref{table:teststorms}, which is the same list of storm events as used in \citet{Ji2012}.
 
@@
 
-Figures \ref{fig:CompareMae} and \ref{fig:CompareCC} show how the mean absolute error and coefficient of correlation as calculated on the validation set storm events of table \ref{table:validationstorms}, vary with increasing model order for GP-AR and GP-ARX.
-The results are represented as boxplots, in which a rectangle is drawn to represent the second and third quartiles, with a vertical line inside to indicate the median value while outlying points are shown as dots.
+Figures \ref{fig:CompareMae} and \ref{fig:CompareCC} show how the mean absolute error and coefficient of correlation as calculated on the validation set storm events of Table \ref{table:validationstorms}, vary with increasing model order for GP-AR and GP-ARX.
+The results are represented as boxplots, in which a rectangle is drawn to represent the first and third quartiles, with a vertical line inside to indicate the median value while outlying points are shown as dots.
 In both cases, the predictive performance first improves and then stagnates or worsens with increasing model order.
@@
 
-After choosing the best performing GP-AR and GP-ARX models, we calculate their performance on the test set of table \ref{table:teststorms}.
-The results of these model evaluations are summarized in table \ref{table:results}, the GP-AR and GP-ARX models improve upon the performance of the \emph{persistence model}.
+After choosing the best performing GP-AR and GP-ARX models, we calculate their performance on the test set of Table \ref{table:teststorms}.
+The results of these model evaluations are summarized in Table \ref{table:results}, the GP-AR and GP-ARX models improve upon the performance of the \emph{persistence model}.
 
@@
 Leveraging the strengths of the Bayesian approach, they are able to learn robust predictors from data.
-If one compares the training sets used by all the models, one can appreciate that the models presented here need relatively small training and validations sets: the training set contains 243 instances, while the validation set contains 782 instances.
+If one considers the size of the data used in our study, one can appreciate that the models presented here need relatively small training and validations sets: the training set contains 243 instances, while the validation set contains 782 instances.
 

### 2017-03-16 agutmpl.tex
Penultimate changes 2

@@
 Figures \ref{fig:CompareMae} and \ref{fig:CompareCC} show how the mean absolute error and coefficient of correlation as calculated on the validation set storm events of Table \ref{table:validationstorms}, vary with increasing model order for GP-AR and GP-ARX.
-The results are represented as boxplots, in which a rectangle is drawn to represent the first and third quartiles, with a vertical line inside to indicate the median value while outlying points are shown as dots.
+The results are represented as box and whisker plots, in which a rectangle is drawn to represent the first and third quartiles, with a horizontal line inside to indicate the median value, outlying points are shown as dots while the whiskers indicate the smallest and largest non-outliers.
 In both cases, the predictive performance first improves and then stagnates or worsens with increasing model order.
@@
 \caption{Mean Absolute Error on validation set storms vs model order for GP-AR and GP-ARX.
-\\ \textbf{Key}: Rectangle borders represent the second and third quartiles, with a horizontal line inside to indicate the median value while outlying points are shown as dots}
+\\ \textbf{Key}: Rectangle borders represent the first and third quartiles, with a horizontal line inside to indicate the median value, outlying points are shown as dots and whiskers indicate the smallest and largest non-outliers}
 \label{fig:CompareMae}
@@
 \noindent\includegraphics[width=\textwidth]{Compare-cc.png}
-\caption{Coefficient of Correlation on validation set storms vs model order for GP-AR and GP-ARX \\ \textbf{Key}: Rectangle borders represent the second and third quartiles, with a horizontal line inside to indicate the median value while outlying points are shown as dots}
+\caption{Coefficient of Correlation on validation set storms vs model order for GP-AR and GP-ARX \\ \textbf{Key}: Rectangle borders represent the first and third quartiles, with a horizontal line inside to indicate the median value, outlying points are shown as dots and whiskers indicate the smallest and largest non-outliers}
 \label{fig:CompareCC}
@@
 \noindent\includegraphics[width=\textwidth]{Compare-mae-arx.png}
-\caption{Mean Absolute Error on validation set storms vs model order for GP-AR and GP-ARX for \emph{CSA}, \emph{GS} and \emph{ML} model selection routines \\ Rectangle borders represent the second and third quartiles, with a horizontal line inside to indicate the median value while outlying points are shown as dots}
+\caption{Mean Absolute Error on validation set storms vs model order for GP-AR and GP-ARX for \emph{CSA}, \emph{GS} and \emph{ML} model selection routines \\ \textbf{Key}: Rectangle borders represent the first and third quartiles, with a horizontal line inside to indicate the median value, outlying points are shown as dots and whiskers indicate the smallest and largest non-outliers}
 \label{fig:CompareMaeARX}
@@
 \noindent\includegraphics[width=\textwidth]{Compare-cc-arx.png}
-\caption{Coefficient of Correlation on validation set storms vs model order for GP-AR and GP-ARX for \emph{CSA}, \emph{GS} and \emph{ML} model selection routines \\ Rectangle borders represent the second and third quartiles, with a horizontal line inside to indicate the median value while outlying points are shown as dots}
+\caption{Coefficient of Correlation on validation set storms vs model order for GP-AR and GP-ARX for \emph{CSA}, \emph{GS} and \emph{ML} model selection routines \\ \textbf{Key}: Rectangle borders represent the first and third quartiles, with a horizontal line inside to indicate the median value, outlying points are shown as dots and whiskers indicate the smallest and largest non-outliers}
 \label{fig:CompareCCARX}

### 2017-05-22 revised-draft.tex
Updates from ShareLaTeX

--- 2017-03-16 agutmpl.tex
@@
 One such work described in \citet{McPherron:2013} involves identification of high speed solar wind streams using the WSA model (see \citet{WSAModel}), using predictions of high speed streams to construct ensembles of $Dst$ trajectories which yield the quartiles of $Dst$ time series.
+A simple way to construct error bars on the predictions of forecasting models is by using the so called \textit{past cast} performance i.e.
+by calculating the standard deviations of the predictions generated by the model on a hold out data set.
+One limitation of such an approach is that the variance of the model predictions, once computed, remains static and does not adapt according to the inputs provided to the model.
+This may lead to overestimation or underestimation of the uncertainty around a models predictions depending on the prevelant geo-magnetic conditions and the data set used to calculate the \textit{past cast} model performance.
 

### 2017-05-26 revised-draft.tex
Added section on Akasofu epsilon

@@
 
-    \item $Dst$: A continuous hourly index which gives a measure of the weakening or strengthening of the Earth's equatorial magnetic field due to the weakening or strengthening of the ring currents and the geomagnetic storms \citep{DesslerAndParker}.
+    \item $Dst$: A continuous hourly index which gives a measure of the weakening or strengthening of the Earth's equatorial magnetic field due to particle injection in the magnetosphere.
+Particle injection has a number of sources such as, weakening or strengthening of the ring currents and the geomagnetic storms \citep{DesslerAndParker}, near Earth cross tail current \citep{ganushkina2004long} and \citep{angeo-28-123-2010}, partial ring current \citep{JGRA:JGRA15878}, substorm current wedge \citep{JGRA:JGRA15211}, magnetopause current, etc .
 \end{enumerate}
@@
 The parameters of the so called NARMAX OLS-ERR model are calculated by solving the \emph{ordinary least squares} (OLS) problem arising from a quadratic objective function.
+It must be noted that the NARMAX methodology is not limited to polynomial functions, rather any set of basis function expansions can be used with it, such as but not limited to radial basis functions, wavelets etc \citep{doi:10.1080/00207720600903011}, \citep{JGRA:JGRA17327}.
 The reader may refer to \citet{billings2013nonlinear} for a detailed exposition of the NARMAX methodology.
@@
 Auto-regressive models can be augmented by including exogenous quantities in the inputs $\mathbf{x}_{t-1}$ at each time step, in order to improve predictive accuracy.
+While modeling $Dst$ using the OMNI data, one must choose which solar wind quantities to include in the exogenous inputs of the predictive model.
+This choice is not straight forward and eventually requires a compromise between including important solar wind quantities and keeping the input space managable in the interest of simplicity.
+
+\subsubsection{Choice of Solar Wind Inputs}
+
 $Dst$ gives a measure of ring currents, which are modulated by plasma sheet particle injections into the inner magnetosphere during sub-storms.
@@
 An increase in the solar wind electric field, $V_{sw}B_z$, can increase the dawn-dusk electric field in the magnetotail, which in turn determines the amount of plasma sheet particle that move to the inner magnetosphere \citet{Friedel2001}.
-Therefore, our exogenous parameters consist of solar wind velocity $V_{sw}$ and IMF $B_z$.
 
-In this model we choose distinct time lags $p$, $p_{v}$ and $p_{b}$ for $Dst$, $V$ and $B_z$ respectively.
+Apart from $V$ and $B_z$, other quantities which have been shown to correlate with geomagnetic activity are solar wind dynamic pressure $P$, clock angle $tan \theta = \frac{B_y}{B_z}$, Akasofu $\epsilon$ \citep{1986AkasofuE} and solar wind magnetosphere coupling functions \citep{JGRA:JGRA21451}.
+
+Although solar wind magnetospheric coupling functions have been shown to have correlation with geomagnetic indices, they are expressible in terms of $V$ and $B_z$, and hence we do not need to include them as explicit inputs to the model.
+\emph{Gaussian Process} models derive their strength from automatic feature construction achieved by the covariance functions (interested readers may refer to chapters 6 and 7  in \cite{Rasmussen:2005:GPM:1162254}).
+Therefore as long as coupling functions can be approximated in the \emph{eigenspace} of the covariance function we need not make them explicit in the input features.
+
+Therefore, our exogenous parameters consist of solar wind velocity $V_{sw}$ and IMF $B_z$. In this model we choose distinct time lags $p$, $p_{v}$ and $p_{b}$ for $Dst$, $V$ and $B_z$ respectively.
 

### 2017-05-29 revised-draft.tex
Added new references

@@
 The parameters of the so called NARMAX OLS-ERR model are calculated by solving the \emph{ordinary least squares} (OLS) problem arising from a quadratic objective function.
-It must be noted that the NARMAX methodology is not limited to polynomial functions, rather any set of basis function expansions can be used with it, such as but not limited to radial basis functions, wavelets etc \citep{doi:10.1080/00207720600903011}, \citep{JGRA:JGRA17327}.
+It must be noted that the NARMAX methodology is not limited to polynomial functions, rather any set of basis function expansions can be used with it, such as radial basis functions, wavelets etc \citep{doi:10.1080/00207720600903011}, \citep{JGRA:JGRA17327}.
 The reader may refer to \citet{billings2013nonlinear} for a detailed exposition of the NARMAX methodology.
@@
 \citet{pallocchia:hal-00318011} proposed a \emph{neural network} model called EDDA to predict $Dst$ using only the IMF data.
+
+%Local neurofuzzy modelling and singular spectrum analysis.
+Apart from the NARMAX and neural network approaches, fuzzy methods have also been applied for $Dst$ prediction, \citet{SWE:SWE146} and \citet{Sharifi2006} outline the application of \emph{Local Neurofuzzy} models for one hour and two hour predictions of $Dst$ respectively.
+Local neuro-fuzzy models reduce the input space into a number of regions each with its own expert predictor.
+The combined model predicts $Dst$ for a new point as a linear combinations of the predictions from each expert weighted by a fuzzy score signifying the importance of each model for the provided input.
+For improving predictive performance of two hour $Dst$ forecasts the authors use \emph{singular spectrum analysis} SSA \citet{loskutov2001testing}, \citet{loskutov2001study}.
 
@@
 
-Although solar wind magnetospheric coupling functions have been shown to have correlation with geomagnetic indices, they are expressible in terms of $V$ and $B_z$, and hence we do not need to include them as explicit inputs to the model.
+Although solar wind magnetospheric coupling functions have been shown to have correlation with geomagnetic indices, they are expressed in terms of $V$ and $B_z$, and hence we do not include them as explicit inputs to the model.
 \emph{Gaussian Process} models derive their strength from automatic feature construction achieved by the covariance functions (interested readers may refer to chapters 6 and 7  in \cite{Rasmussen:2005:GPM:1162254}).
-Therefore as long as coupling functions can be approximated in the \emph{eigenspace} of the covariance function we need not make them explicit in the input features.
+As long as coupling functions can be approximated in the \emph{eigenspace} of the covariance function we need not make them explicit in the input features.
 
@@
 
+It is an important question as to how unaccounted inputs such as solar wind dynamic pressure $P$ and clock angle $\theta$ affect the structure of the GP-ARX model.
+From a model selection perspective, these unaccounted inputs should lead to higher values of the noise covariance.
+In the specific case of solar wind dynamic pressure, it is calculated as a product of the plasma density and the solar wind speed making it highly correlated with the solar wind speed, as a result the GP-ARX model can infer a large portion of the information content from the solar wind speed itself.
+With respect to the clock angle, it must be noted that coupling functions such as Akasofu $\epsilon$ generally contain powers of $sin \theta$ bounding the effect of clock angle to an absolute magnitude of $1$, hence we do not expect these unaccounted inputs to greatly improve the predictive capabilities of the GP-ARX model.
+
 \subsection{Choice of Mean Function}
@@
 If there is no prior knowledge about the function to be approximated, \citet{Rasmussen:2005:GPM:1162254} state that it is perfectly reasonable to choose $m(\mathbf{x} = 0)$ as the mean function, as long as the target values are normalized.
-In the case of the $Dst$ time series, it is known that the so called \emph{persistence model} $\hat{D}st(t) = Dst(t-1)$ performs quite well in the context of OSA prediction.
-We therefore choose the \emph{persistence model} as the mean function in our OSA Dst models.
+In the case of the $Dst$ time series, it is known that the so called \emph{persistence model} $\hat{D}st(t) = Dst(t-1)$ has high correlation with \emph{OSA} $Dst$.
+Due to its simplicity, we choose the \emph{persistence model} as the prior mean function in our OSA Dst models.
+
+The \emph{persistence model} can be described as \emph{Markovian} prediction mechanism, when it is chosen as the prior mean of the GP-AR and GP-ARX model, it is indeed the case that the prior probability distribution of $Dst(t)$ is Gaussian with a strong \emph{Markovian} behavior $P(Dst(t)|\mathbf{x}_t) \sim \mathcal{N}(Dst(t-1), \sqrt{K_{osa}(\mathbf{x}_t, \mathbf{x}_t)})$, but the posterior predictive distribution of $Dst(t)$ conditional on the model training data (given in equation \ref{eq:posteriormean}) is non-\emph{Markovian} due its dependence on the term denoted by $\mathbf{K}_{*}$ which contains kernel values computed between the test data and training data features.
+Thus the GP-AR and GP-ARX models when used conditional on training data are non-\emph{Markovian} predictive models.
 
@@
       \item \emph{Persistence} model plays an important role in the model building and evaluation process in the context of \emph{one step ahead} prediction of the $Dst$ index.
-It is clear that the persistence behavior in the $Dst$ values is very strong i.e.
-the trivial predictive model $\hat{Dst}(t) = Dst(t-1)$ gives excellent performance according to the metrics chosen.
+Although it is not a robust predictor for the onset of intense geomagnetic storms, the \emph{persistence model} performs well on classical error metrics such as \emph{root mean square error} and such.
+From the considerations above, it is quite evident that classical performance metrics are not adequate for model evaluation, nevertheless in space weather literature, metrics such as \emph{RMSE} are very commonly used to compare predictive performance of models.
+Although not the research focus of this study, we note that there exists a need for the formulation of more informative performance metrics for measurement of predictive performance of geomagnetic predictive models.
 

### 2017-05-30 revised-draft.tex
Revised manuscript

@@
 From the validation results, we can chose the model order which yields the best performance, for GP-AR it is $p_t = 6$ while for GP-ARX it is $p_t = 11$.
-Further examination of the validation results shows that in the scheme $p_t = 11$ choosing $p = 7, p_v = 1, p_b = 3$ gives superior results.
+Further examination of the validation results shows that in the scheme $p_t = 11$ choosing $p = 6, p_v = 1, p_b = 3$ gives superior results.
 
@@
 \noindent\includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm43.png}
-\caption{OSA Predictions with $\pm \sigma$ error bars for event: 2003/06/17 19:00 to 2003/06/19 03:00}
+\caption{OSA Predictions with $\pm \sigma$ error bars for event: 2003/06/17 to 2003/06/19}
 \label{fig:ComparePred1}
@@
 \begin{figure}
-\noindent\includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm8.png}
-\caption{OSA Predictions with $\pm \sigma$ error bars for event: 1998/11/13 00:00 to 1998/11/15 04:00}
+\noindent\includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm16.png}
+\caption{OSA Predictions with $\pm \sigma$ error bars for event: 2012/03/08 to 2012/03/10}
 \label{fig:ComparePred2}
@@
 \begin{figure}
-\noindent\includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm9.png}
-\caption{OSA Predictions with $\pm \sigma$ error bars for event: 1999/01/13 16:00 to 1999/01/14 20:00}
+\noindent\includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm46.png}
+\caption{OSA Predictions with $\pm \sigma$ error bars for event: 2003/11/20 to 2003/11/22}
 \label{fig:ComparePred3}
@@
 Model & Mean Absolute Error & Root Mean Square Error & Coefficient of Correlation\\ \hline
-GP-ARX & 7.252 & 11.93 & 0.972\\
+GP-ARX & 7.219 & 11.88 & 0.972\\
 GP-AR & 8.37 & 14.04 & 0.963\\

### 2017-05-30 response-reviewer1.tex
Response Reviewer 1: [WIP]

\documentclass{article}
\begin{document}

\title{Response Reviewer 1}
\maketitle

\section*{Draft Comments \& Revisions}

\begin{enumerate}
\item{
  \fbox{%
    \parbox{\textwidth}{%
line 31: Dst: A continuous hourly index which gives a measure of the weakening or strengthening of the Earth's equatorial magnetic field due to the weakening or strengthening of the ring currents and the geomagnetic storms. \\
line 204: Dst gives a measure of ring currents. 
\\
\textbf{Comment}: Authors should add that ring current is not the only current that contributes to the Dst evolution. 
The authors should mention the contribution of near -Earth cross-tail current (e.g. Ganushkina, et al., Ann. Geophys. 22, 1317-1334, 2004; Ganushkina, et al.. Ann. Geophys. 28, 123- 140, 2010); partial ring current 
(e.g. Liemohn et al. J. Geophys. Res. 106, 10883-10904, 2001), substorm current wedge (e.g. Munsami, J. Geophys. Res. 105(A12), 27833-27840, 2000); magnetopause current. }
}

\textbf{Response}: Changes made in lines 32-36 in the file \texttt{changes.pdf}

}

\item{

\fbox{%
\parbox{\textwidth}{%
line 52-53: "The NARMAX methodology builds models by constructing polynomial expansions of inputs and determines the best combinations of monomials to include in the refined model by using a criterion called the error reduction ratio (ERR)."

\textbf{Comment}: NARMAX similar to NARX modelling is not limited to the polynomial based approach. It could employ wide range of sets of functions that form basis e.g. radial basis functions, wavelets etc. (e.g.: Wei et al., Wavelet based non-parametric NARX models for nonlinear input-output system identification; International Journal of Systems Science, Volume: 37, Issue: 15, Pages: 1089-1096, DOI: 10.1080/00207720600903011, Published: DEC 15 2006; Wei et al., Prediction of the Dst index using multiresolution wavelet models; JGR-SP, 109, A7, DOI: 10.1029/2003JA010332, 2004). Please modify the statement about NARMAX.

}
}

\textbf{Response}: Changes made in lines 59-62 in the file \texttt{changes.pdf}

}

\item{

\fbox{%
\parbox{\textwidth}{%
line 122: "Since we have noisy measurements of f over the training data, we add the noise variance σ2"

\textbf{Comment}: V and Bz are used in the current manuscript for GP-ARX model. It is well known that P dynamic pressure affects the Dst. Please discuss if the un-accounted inputs like P can affect statistical properties of "noise" and if these un-accounted inputs can affect the structure of the model.

}
}

\textbf{Response}: Revisions made in lines 239-249 in the file \texttt{changes.pdf}

}

\item{

\fbox{%
\parbox{\textwidth}{%
"Therefore, our exogenous parameters consist of solar wind velocity Vsw and IMF Bz."

\textbf{Comment}: There were many previously published papers that tried to identify the coupling functions between the solar wind and the Dst index. Since the introduction of Acasofu "epsilon" parameter many coupling functions include also "clock angle". Please explain in more detail why it was decided not to include "clock angle" into the set of parameters.

}
}

\textbf{Response}: Revisions made in lines 239-249 in the file \texttt{changes.pdf}

}

\item{

\fbox{%
\parbox{\textwidth}{%
Line 219: "It is known that the so called persistence model Dˆ st(t) = Dst(t − 1) performs quite well in the context of OSA prediction"

\textbf{Comment}: I think the major aim of various Dst forecasting tools is to predict in advance sever geomagnetic storms. Persistence model will fail exactly at the moment when accurate prediction is required, one step before the initiation of major geomagnetic storms. If some statistical parameters show that the persistence model performs quite well, thats mean that these parameters are not adequate in the assessment of forecasting tools and some other parameters should be used. Please modify the phrase above.

}
}

\textbf{Response}: Revisions made in lines 268-270 and 340-347 in the file \texttt{changes.pdf}

}

\item{

\fbox{%
\parbox{\textwidth}{%

\textbf{Comment}: Figures 5,6,7. Please expand the time interval shown. The figure should include the reasonable period of quiet time before and after the geomagnetic storm.

}
}

\textbf{Response}: Changes made for the figures 5-7

}

\end{enumerate}

\end{document}

### 2017-05-31 response-reviewer1.tex
Response Reviewers added

@@
 \begin{enumerate}
+
+\item{
+
+\fbox{%
+    \parbox{\textwidth}{%
+\textbf{Comment}: The authors state the proposed methodology allows to construct error bars on the predictions.
+I think this can not be considered as a significant advantage because the past-cast performance for any forecasting tool can be used to construct statistical error bars for the prediction.}
+}
+
+\textbf{Response}: Changes made in lines 91-97 in the file \texttt{changes.pdf}
+
+}
+
 \item{
@@
 
-\textbf{Response}: Revisions made in lines 239-249 in the file \texttt{changes.pdf}
+\textbf{Response}: Revisions made in lines 229-233, 243-253 and 256-265 in the file \texttt{changes.pdf}
 
@@
 
-\textbf{Response}: Revisions made in lines 239-249 in the file \texttt{changes.pdf}
+\textbf{Response}: Revisions made in lines 243-253 in the file \texttt{changes.pdf}
 
@@
 
-\textbf{Response}: Revisions made in lines 268-270 and 340-347 in the file \texttt{changes.pdf}
+\textbf{Response}: Revisions made in lines 272-273 and 344-351 in the file \texttt{changes.pdf}
 

### 2017-05-31 response-reviewer2.tex
Response Reviewers added

\documentclass{article}
\begin{document}

\title{Response Reviewer 2}
\maketitle

\section*{Draft Comments \& Revisions}

\begin{enumerate}

\item{

\fbox{%
    \parbox{\textwidth}{%
\textbf{Comment}:

The authors must be care to review completely the researches from classical to intelligent methods for prediction of Dst, Kp. For example some of references are missing is: 
(a)- Javad Sharifie, Caro Lucas, and Babak N. Araabi, "Locally linear neurofuzzy modeling and prediction 
of geomagnetic disturbances based on solar wind conditions", Space Weather, Vol 4, 2006 
(b)- Javad Sharifi, Babak N. Araabi, and Caro Lucas, "Multi-step prediction of Dst index using singular spectrum analysis and locally linear neurofuzzy modeling", Earth Planets Space, 58, 331-341, 2006 
(c)- Loskutov, A., I. A. Istomin, K. M. Kuzanyan, and O. L. Kotlyarov, Testing and forecasting the time series of the solar activity by singular spectrum analysis, Nonlin. Phenomena in Complex Syst., 4(1), 47-57, 2001a. 
(d)- Loskutov, A., I. Istomin, O. Kotlyarov, and K. Kuzanyan, A study of the regularities in Solar magnetic activity by singular spectrum analysis, Astronomy Letters, 27(11), 745-753, 2001b 
And some of other which especially used for Dst predition from one step to multi-step.

}
}

\textbf{Response}: Revisions made in lines 74-85 in the file \texttt{changes.pdf}

}

\item{

\fbox{%
    \parbox{\textwidth}{%
\textbf{Comment}: To select best number of time lags of input regressor vector, you must use structure identification. To do this, an error criteria must be obtain based on number of time lags of each physical component. The error criteria has a minimum which is appropriate number of lagged time series.

}
}

\textbf{Response}: In section 4, we have specified that the model lags for each input component are chosen based on the predictive performance of each generated model on the validation data set. The storm events which comprise the validation data set are specified in table 4 of the manuscript. It is based on this method that we choose the time lags for each input. Care has been taken that no events of the validation data set overlap with the training or test data to ensure unbiased model lag selection.

}

\item{

\fbox{%
    \parbox{\textwidth}{%
\textbf{Comment}: In this paper, even for one step, has one-step time delay, which means that prediction is near Markov (The last observed data is the prediction of next time step), it is obvious from figures 5-7. Then the prediction results is useless. To enhance this, some corrections is in front of authors: they may correct the method to non-Markovian OSA, or use the hybrid prediction methods for example combination of spectrum analysis with OSA or any other hybrid methods.

}
}

\textbf{Response}: The GP-AR and GP-ARX models are strictly speaking non-Markovian models which we have explained in lines 274-281 in the file \texttt{changes.pdf}. Although many predictive models for $Dst$ are non-Markovian in formulation, due to strong \emph{persistence} behavior in the OMNI data, the Markovian component is always a dominant portion of all $Dst$ predictions.

This can be observed in figures 7, 9 in Javad Sharifi, Babak N. Araabi, and Caro Lucas, "Multi-step prediction of Dst index using singular spectrum analysis and locally linear neurofuzzy modeling", Earth Planets Space, 58, 331-341, 2006.

The NARMAX Dst OSA model as outlined in equation 2 of R. J. Boynton, M. A. Balikhin, S. A. Billings, A. S. Sharma, and O. A. Amariutei, "Data derived NARMAX Dst model", has a strong persistence/Markovian component ($0.8335 Dst(t-1)$) which is more significant as compared to the other terms of the model expansion.

We, the authors argue that this is an artifact of the OMNI data and not inductive bias of the $Dst$ models in question.

}

\item{

\fbox{%
    \parbox{\textwidth}{%
\textbf{Comment}: For enhanced method which author must develop, several extreme geomagnetic storms must be used to validate the method. In this storms the Dst index must be below of -300nT, i.e. Dst < -300nT for example extreme geomagnetic storm of 1989 (Dst goes to about -600nT) and etc. in this way, the author can postulate that have contribution for geomagnetic storm prediction.
}
}

\textbf{Response}: With respect to the storm of 1989, there is very little solar wind $V$ and IMF $B_z$ data available for that particular event, hence it was not possible to calculate predictions for it. The authors would however, like to bring to attention event numbers 16 and 43 in table 5 of the manuscript. These events are a part of the test set used to evaluate the GP-AR and GP-ARX models. We have shown predictions for both in tables 6 and 7 of the revised manuscript.

}

\end{enumerate}
\end{document}

### 2017-05-31 revised-draft.tex
Response Reviewers added

@@
 The combined model predicts $Dst$ for a new point as a linear combinations of the predictions from each expert weighted by a fuzzy score signifying the importance of each model for the provided input.
-For improving predictive performance of two hour $Dst$ forecasts the authors use \emph{singular spectrum analysis} SSA \citet{loskutov2001testing}, \citet{loskutov2001study}.
+For improving predictive performance of two hour $Dst$ forecasts in \citet{Sharifi2006}, the authors use \emph{singular spectrum analysis} (SSA).
+Singular spectrum analysis consists of extracting orthogonal components from a lagged time series, it is equivalent to \emph{principal component analysis} (PCA) which is quite extensively used in the machine learning community.
+\citet{loskutov2001testing} and \citet{loskutov2001study} provide a good background to the theory and application of SSA to geomagnetic time series.
 
@@
 One such work described in \citet{McPherron:2013} involves identification of high speed solar wind streams using the WSA model (see \citet{WSAModel}), using predictions of high speed streams to construct ensembles of $Dst$ trajectories which yield the quartiles of $Dst$ time series.
+
 A simple way to construct error bars on the predictions of forecasting models is by using the so called \textit{past cast} performance i.e.
@@
 
-From the validation results, we can chose the model order which yields the best performance, for GP-AR it is $p_t = 6$ while for GP-ARX it is $p_t = 11$.
-Further examination of the validation results shows that in the scheme $p_t = 11$ choosing $p = 6, p_v = 1, p_b = 3$ gives superior results.
+From the validation results, we choose the model order which yields the best RMSE performance, for GP-AR it is $p_t = 6$ while for GP-ARX it is $p = 6, p_v = 1, p_b = 3$.
 
@@
 15 & 2000/05/24 & 0100 & 2000/05/25 & 2000 & -147 \\
-16 & 2000/08/10 & 2000 & 2000/08/11 & 1800 & -106 \\
+16 & 2000/08/10 & 2000 & 2000/08/11 & 1800 & -230 \\
 17 & 2000/08/12 & 0200 & 2000/08/13 & 1700 & -235 \\

### 2017-05-31 letter-editor.tex
Added letter to editor tex file

\documentclass{letter}
\begin{document}

To \\
Editorial board, AGU Space Weather,\\

\hfill

\end{document}

### 2017-05-31 response-reviewer1.tex
Updates to responses
Mods to reviewer response

@@
 \maketitle
+
+The authors would like to thank the reviewer for the various comments and suggestions which greatly helped in improving the quality of the manuscript.
+In the proceeding section they will find the revisions/responses for each concern outlined.
 
@@
 
-\textbf{Response}: Changes made in lines 91-97 in the file \texttt{changes.pdf}
+\textbf{Response}: We do not agree with this assertion.
+To further elaborate the advantage of generating error bars using the Gaussian Process approach we have made changes in lines 91-97 in the file \texttt{changes.pdf}
 

### 2017-06-01 response-reviewer1.tex
Improvements response 1

@@
 
-\textbf{Response}: We do not agree with this assertion.
+\textbf{Response}: We do not agree with this assertion because the error bars calculated from a past cast performance are static with respect to the model inputs.
+That is not the case with error bars generated from the GP-AR and GP-ARX models.
+From figures 5-7 in the manuscript it can be seen that the error bars on a prediction depend on the inputs provided, this dependence is quantified via the covariance function.
 To further elaborate the advantage of generating error bars using the Gaussian Process approach we have made changes in lines 91-97 in the file \texttt{changes.pdf}
@@
 
-\textbf{Response}: Changes made in lines 32-36 in the file \texttt{changes.pdf}
+\textbf{Response}: We agree with the recommendations. Changes made in lines 32-36 in the file \texttt{changes.pdf}
 
@@
 
-\textbf{Response}: Changes made in lines 59-62 in the file \texttt{changes.pdf}
+\textbf{Response}: We agree with the recommendations. Changes made in lines 59-62 in the file \texttt{changes.pdf}
 
@@
 
-\textbf{Response}: Revisions made in lines 229-233, 243-253 and 256-265 in the file \texttt{changes.pdf}
+\textbf{Response}: We appreciate this comment as it raises important modelling questions.
+Our general understanding is that unaccounted solar wind parameters would increase the magnitude of the noise component in the model, on the other hand we do not have strong reasons to think that it will drastically change the structure of the model with respect to parameters such as the autoregressive lag of $Dst$ and the exogenous inputs.
+We feel this because the solar wind speed and IMF $B_z$ contain a large amount of statistical information with regards to the dynamics of $Dst$.
+Also worth noting is that the solar wind dynamic pressure in the OMNI data is calculated as a product of plasma density and square of solar wind speed.
+Revisions have been made along these ideas in lines 229-233, 243-253 and 256-265 in the file \texttt{changes.pdf}
 
@@
 
-\textbf{Response}: Revisions made in lines 243-253 in the file \texttt{changes.pdf}
+\textbf{Response}: With regards to coupling functions, they can be approximated in the GP-ARX model as long as the eigen-space of the chosen covariance allows functional expressions which resemble common solar wind magnetosphere coupling functions.
+It is also the case that a large portion of the magnitude of these coupling functions consists of the solar wind parameters already included in the GP-ARX model.
+Since the clock angle enters these coupling functions through the term $sin^4 \theta$, its magnitude and impact is bounded.
+Therefore we felt it was not imperative to not include it in the model exogenous inputs.
+Revisions made in lines 243-253 in the file \texttt{changes.pdf}
 
@@
 
-\textbf{Response}: Revisions made in lines 272-273 and 344-351 in the file \texttt{changes.pdf}
+\textbf{Response}: We agree with this comment, the persistence model will fail one moment before a storm, also it is the need of the hour for more informative performance metrics for $Dst$ prediction.
+Since it was not the aim of this research to suggest new performance metrics, we chose to use the ones which are widely used in publications so far.
+Revisions made in lines 272-273 and 344-351 in the file \texttt{changes.pdf}
 

### 2017-06-01 response-reviewer2.tex
Updates to responses
Improvements response 2

@@
 
-\textbf{Response}: Revisions made in lines 74-85 in the file \texttt{changes.pdf}
+\textbf{Response}: We thank the reviewer for pointing out these publications. Revisions made in lines 74-85 in the file \texttt{changes.pdf}
 

### 2017-06-06 response-reviewers_merged.tex
Changes Enrico

\documentclass{article}

\usepackage[parfill]{parskip}

\begin{document}

\title{Reply to Reviewers}
\maketitle

The authors would like to thank the reviewers for the various comments and suggestions which greatly helped in improving the quality of the manuscript. 
In the proceeding section they will find the revisions/responses for each concern outlined.

\section*{Reviewer 1}

\textit{
\textbf{Comment 1)}: The authors state the proposed methodology allows to construct error bars on the predictions. I think this can not be considered as a significant advantage because the past-cast performance for any forecasting tool can be used to construct statistical error bars for the prediction.}\\
\\
\textbf{Response 1)}: We do not agree with this assertion because the error bars calculated from a past-cast performance are static with respect to the model inputs. That is not the case with error bars generated from the GP-AR and GP-ARX models. From figures 5-7 in the manuscript it can be seen that the error bars on a prediction depend on the inputs provided, this dependence is quantified via the covariance function. To further elaborate the advantage of generating error bars using the Gaussian Process approach we have added the following paragraph (lines 91-97).\\
\\
\fbox{\parbox{\textwidth}{
A simple way to construct error bars on the predictions of forecasting models is by using the so called \textit{past cast} performance i.e. by calculating the standard deviations of the predictions generated by the model on a hold out data set. One limitation of such an approach is that the variance of the model predictions is computed once and for all. It does not adapt according to the inputs provided to the model. This may lead to overestimation or underestimation of the uncertainty around a given prediction, depending on the prevelant geo-magnetic conditions and the data set used to calculate the \textit{past cast} model performance.}
}\\
\\

{\it \textbf{Comment 2)}: Authors should add that ring current is not the only current that contributes to the Dst evolution. 
The authors should mention the contribution of near -Earth cross-tail current (e.g. Ganushkina, et al., Ann. Geophys. 22, 1317-1334, 2004; Ganushkina, et al.. Ann. Geophys. 28, 123- 140, 2010); partial ring current 
(e.g. Liemohn et al. J. Geophys. Res. 106, 10883-10904, 2001), substorm current wedge (e.g. Munsami, J. Geophys. Res. 105(A12), 27833-27840, 2000); magnetopause current. }\\

\textbf{Response 2)}: We agree with the recommendations. We have added those references.\\
\\

\textit{
\textbf{Comment 3)}: NARMAX similar to NARX modelling is not limited to the polynomial based approach. It could employ wide range of sets of functions that form basis e.g. radial basis functions, wavelets etc. (e.g.: Wei et al., Wavelet based non-parametric NARX models for nonlinear input-output system identification; International Journal of Systems Science, Volume: 37, Issue: 15, Pages: 1089-1096, DOI: 10.1080/00207720600903011, Published: DEC 15 2006; Wei et al., Prediction of the Dst index using multiresolution wavelet models; JGR-SP, 109, A7, DOI: 10.1029/2003JA010332, 2004). Please modify the statement about NARMAX. 
}\\

\textbf{Response 3)}: We agree with the recommendations. The following changes have been made (lines 59-62)\\
\\
\fbox{%
\parbox{\textwidth}{It must be noted that the NARMAX methodology is not limited to polynomial functions, rather any set of basis function expansions can be used with it, such as radial basis functions, wavelets etc [Wei et al. 2006; 2004]	.}}\\
\\

\textit{
\textbf{Comment 4)}: V and Bz are used in the current manuscript for GP-ARX model. It is well known that P dynamic pressure affects the Dst. Please discuss if the un-accounted inputs like P can affect statistical properties of "noise" and if these un-accounted inputs can affect the structure of the model.}\\

\textbf{Response 4)}: We appreciate this comment as it raises important modelling questions. Our general understanding is that unaccounted solar wind parameters would increase the magnitude of the noise component in the model, on the other hand we do not have strong reasons to think that it will drastically change the structure of the model with respect to parameters such as the autoregressive lag of $Dst$ and the exogenous inputs. We feel this because the solar wind speed and IMF $B_z$ contain a large amount of statistical information with regards to the dynamics of $Dst$. Also worth noting is that the solar wind dynamic pressure in the OMNI data is calculated as a product of plasma density and square of solar wind speed. Revisions have been made along these ideas. \\

Lines 229-233:\\
\fbox{%
\parbox{\textwidth}{While modeling $Dst$ using the OMNI data, one must choose which solar wind quantities to include in the exogenous inputs of the predictive model. This choice is not straight forward and eventually requires a compromise between including important solar wind quantities and keeping the input space managable in the interest of simplicity.
}}\\

Lines 256-265:\\
\fbox{%
\parbox{\textwidth}{It is an important question as to how unaccounted inputs such as solar wind dynamic pressure $P$ and clock angle $\theta$ affect the structure of the GP-ARX model. From a model selection perspective, these unaccounted inputs should lead to higher values of the noise covariance. In the specific case of solar wind dynamic pressure, it is calculated as a product of the plasma density and the solar wind speed making it highly correlated with the solar wind speed, as a result the GP-ARX model can infer a large portion of the information content from the solar wind speed itself. With respect to the clock angle, it must be noted that coupling functions such as Akasofu $\epsilon$ generally contain powers of $sin \theta$ bounding the effect of clock angle to an absolute magnitude of $1$, hence we do not expect these unaccounted inputs to greatly improve the predictive capabilities of the GP-ARX model.
 }}\\

\textit{
\textbf{Comment 5)}: There were many previously published papers that tried to identify the coupling functions between the solar wind and the Dst index. Since the introduction of Acasofu "epsilon" parameter many coupling functions include also "clock angle". Please explain in more detail why it was decided not to include "clock angle" into the set of parameters. 
}\\

\textbf{Response 5)}: With regards to coupling functions, they can be approximated in the GP-ARX model as long as the eigen-space of the chosen covariance allows functional expressions which resemble common solar wind magnetosphere coupling functions. It is also the case that a large portion of the magnitude of these coupling functions consists of the solar wind parameters already included in the GP-ARX model. Since the clock angle enters these coupling functions through the term $sin^4 \theta$, its magnitude and impact is bounded. Therefore we felt it was not imperative to not include it in the model exogenous inputs. Revisions made in lines 243-253:\\

\fbox{%
\parbox{\textwidth}{Apart from $V$ and $B_z$, other quantities which have been shown to correlate with geomagnetic activity are solar wind dynamic pressure $P$, clock angle $tan \theta = \frac{B_y}{B_z}$, Akasofu $\epsilon$ [Pudovkin and Semenov, 1986] and solar wind magnetosphere coupling functions [Spencer et al. 2011].}}\\

\textit{
Line 219: "It is known that the so called persistence model Dˆ st(t) = Dst(t − 1) performs quite well in the context of OSA prediction".\\ 
\textbf{Comment 6)}: I think the major aim of various Dst forecasting tools is to predict in advance sever geomagnetic storms. Persistence model will fail exactly at the moment when accurate prediction is required, one step before the initiation of major geomagnetic storms. If some statistical parameters show that the persistence model performs quite well, thats mean that these parameters are not adequate in the assessment of forecasting tools and some other parameters should be used. Please modify the phrase above. 
}\\

\textbf{Response 6)}: We agree with this comment, the persistence model will fail one moment before a storm. Since it was not the aim of this research to suggest new performance metrics, we chose to use the ones which are widely used in publications so far. Revisions made in lines 272-273:\\
\\
\fbox{%
\parbox{\textwidth}{In the case of the $Dst$ time series, it is known that the so called \emph{persistence model} $\hat{D}st(t) = Dst(t-1)$ has high correlation with \emph{OSA} $Dst$. Due to its simplicity, we choose the \emph{persistence model} as the prior mean function in our OSA Dst models.} 
}\\
Lines 344-351:\\ 
\\
\fbox{%
\parbox{\textwidth}{
Although it is not a robust predictor for the onset of intense geomagnetic storms, the \emph{persistence model} performs well on classical error metrics such as \emph{root mean square error} and such. From the considerations above, it is quite evident that classical performance metrics are not adequate for model evaluation, nevertheless in space weather literature, metrics such as \emph{RMSE} are very commonly used to compare predictive performance of models. Although not the research focus of this study, we note that there exists a need for the formulation of more informative performance metrics for measurement of predictive performance of geomagnetic predictive models.}}

\textit{
\textbf{Comment 7)}: Figures 5,6,7. Please expand the time interval shown. The figure should include the reasonable period of quiet time before and after the geomagnetic storm.
}\\

\textbf{Response 7)}: Changes made for the figures 5-7

\newpage
\section*{Reviewer 2}

\textit{
\textbf{Comment 1)}: 
The authors must be care to review completely the researches from classical to intelligent methods for prediction of Dst, Kp. For example some of references are missing is: 
(a)- Javad Sharifie, Caro Lucas, and Babak N. Araabi, "Locally linear neurofuzzy modeling and prediction 
of geomagnetic disturbances based on solar wind conditions", Space Weather, Vol 4, 2006 
(b)- Javad Sharifi, Babak N. Araabi, and Caro Lucas, "Multi-step prediction of Dst index using singular spectrum analysis and locally linear neurofuzzy modeling", Earth Planets Space, 58, 331-341, 2006 
(c)- Loskutov, A., I. A. Istomin, K. M. Kuzanyan, and O. L. Kotlyarov, Testing and forecasting the time series of the solar activity by singular spectrum analysis, Nonlin. Phenomena in Complex Syst., 4(1), 47-57, 2001a. 
(d)- Loskutov, A., I. Istomin, O. Kotlyarov, and K. Kuzanyan, A study of the regularities in Solar magnetic activity by singular spectrum analysis, Astronomy Letters, 27(11), 745-753, 2001b 
And some of other which especially used for Dst predition from one step to multi-step.
}

\textbf{Response 1)}: We thank the reviewer for pointing out these publications. Revisions made in lines 74-85:\\
\\
\fbox{%
\parbox{\textwidth}{
Apart from the NARMAX and neural network approaches, fuzzy methods have also been applied for $Dst$ prediction, Sharifie et al. [2006] outline the application of \emph{Local Neurofuzzy} models for one hour and two hour predictions of $Dst$ respectively. Local neuro-fuzzy models reduce the input space into a number of regions each with its own expert predictor. The combined model predicts $Dst$ for a new point as a linear combinations of the predictions from each expert weighted by a fuzzy score signifying the importance of each model for the provided input. For improving predictive performance of two hour $Dst$ forecasts in Sharifie et al. [2006], the authors use \emph{singular spectrum analysis} (SSA). Singular spectrum analysis consists of extracting orthogonal components from a lagged time series, it is equivalent to \emph{principal component analysis} (PCA) which is quite extensively used in the machine learning community. Luskutov et al. [2001a, 2001b] provide a good background to the theory and application of SSA to geomagnetic time series.}}\\

\textit{
\textbf{Comment 2)}: To select best number of time lags of input regressor vector, you must use structure identification. To do this, an error criteria must be obtain based on number of time lags of each physical component. The error criteria has a minimum which is appropriate number of lagged time series. }\\

\textbf{Response 2)}: In section 4, we have specified that the model lags for each input component are chosen based on the predictive performance of each generated model on the validation data set. The storm events which comprise the validation data set are specified in table 4 of the manuscript. It is based on this method that we choose the time lags for each input. Care has been taken that no events of the validation data set overlap with the training or test data to ensure unbiased model lag selection.\\

\textit{
\textbf{Comment 3)}: In this paper, even for one step, has one-step time delay, which means that prediction is near Markov (The last observed data is the prediction of next time step), it is obvious from figures 5-7. Then the prediction results is useless. To enhance this, some corrections is in front of authors: they may correct the method to non-Markovian OSA, or use the hybrid prediction methods for example combination of spectrum analysis with OSA or any other hybrid methods.
}\\

\textbf{Response 3)}: The GP-AR and GP-ARX models are strictly speaking non-Markovian models which we have explained in lines 274-281.\\
\fbox{%
\parbox{\textwidth}{The \emph{persistence model} can be described as \emph{Markovian} prediction mechanism, when it is chosen as the prior mean of the GP-AR and GP-ARX model, it is indeed the case that the prior probability distribution of $Dst(t)$ is Gaussian with a strong \emph{Markovian} behavior $P(Dst(t)|\mathbf{x}_t) \sim \mathcal{N}(Dst(t-1), \sqrt{K_{osa}(\mathbf{x}_t, \mathbf{x}_t)})$, but the posterior predictive distribution of $Dst(t)$ conditional on the model training data (given in equation 13) is non-\emph{Markovian} due its dependence on the term denoted by $\mathbf{K}_{*}$ which contains kernel values computed between the test data and training data features. Thus the GP-AR and GP-ARX models when used conditional on training data are non-\emph{Markovian} predictive models.}
}\\

Although many predictive models for $Dst$ are non-Markovian in formulation, due to strong \emph{persistence} behavior of the DST index, the Markovian component is always a dominant portion of all $Dst$ predictions.

This can be observed in figures 7, 9 in Javad Sharifi, Babak N. Araabi, and Caro Lucas, "Multi-step prediction of Dst index using singular spectrum analysis and locally linear neurofuzzy modeling", Earth Planets Space, 58, 331-341, 2006.

Moreover, also the NARMAX Dst OSA model as outlined in equation 2 of R. J. Boynton, M. A. Balikhin, S. A. Billings, A. S. Sharma, and O. A. Amariutei, "Data derived NARMAX Dst model", has a strong persistence/Markovian component ($0.8335 Dst(t-1)$) which is more significant as compared to the other terms of the model expansion.

We argue that the persistent behaviour of DST is an instrinsic characteristic of the dataset and not an inductive bias of the $Dst$ models in question.

\textit{
\textbf{Comment 5)}: For enhanced method which author must develop, several extreme geomagnetic storms must be used to validate the method. In this storms the Dst index must be below of -300nT, i.e. Dst $<$ -300nT for example extreme geomagnetic storm of 1989 (Dst goes to about -600nT) and etc. in this way, the author can postulate that have contribution for geomagnetic storm prediction.}\\

\textbf{Response 5)}: With respect to the storm of 1989, there is very little solar wind $V$ and IMF $B_z$ data available for that particular event, hence it was not possible to calculate predictions for it. 
The storms used to evaluate the GP-AR and GP-ARX models are listed on Table 5. Once can notice that some of them are are indeed 'extreme storms' with Dst values below -300.

\end{document}

### 2017-06-06 letter-editor.tex
Changes Enrico

@@
 
-To \\
-Editorial board, AGU Space Weather,\\
+Dear Editor,\\
+
+we have substantially revised the manuscript, which we believe is now suitable for publication.
+Reviewer 1 argued that the presented models 'did not result in any significant advance in the forecast of Dst index'.
+We have deliberately avoided to present a direct comparison with other published methods.
+Although that should be the standard way to present and corroborate new models, we have been discouraged to do so, for a number of reasons.
+Any such direct comparison can always be debated when some codes are not publicly available.
+For instance, the code for the celebrated NARMAX model by the Sheffield group is kept secret.
+Hence, trying to reproduce their model is a very error-prone exercise.
+Instead, we have decided to make all our codes available.
+Moreover, we feel that well-established standard metrics on which to evaluate the performance of models are still missing in this community.
+Finally, a major reason to avoid such comparison is that we agree with Referee 2, that the one-hour ahead is of little practical use, because of its strong persistence behavior.
+Nevertheless, by introducing a new methodology, this manuscript represents an important step towards the formulation of multiple hour ahead predictions, which is our next goal.\\
+
+In any case, for your eyes only we include here a figure showing the comparison of our models with other published models. Our model outperforms all of them.
+
+Regarding reviewer 2, we have done our best to address his/her concerns. However, we would like to point out that the meaning of some comments was rather obscure, due to his/her poor English.
+
+Best regards,
+The authors.
 

### 2017-06-06 revised-draft.tex
Changes Enrico

@@
 by calculating the standard deviations of the predictions generated by the model on a hold out data set.
-One limitation of such an approach is that the variance of the model predictions, once computed, remains static and does not adapt according to the inputs provided to the model.
-This may lead to overestimation or underestimation of the uncertainty around a models predictions depending on the prevelant geo-magnetic conditions and the data set used to calculate the \textit{past cast} model performance.
+One limitation of such an approach is that the variance of the model predictions is computed once and for all.
+It does not adapt according to the inputs provided to the model.
+This may lead to overestimation or underestimation of the uncertainty around a given prediction, depending on the prevelant geo-magnetic conditions and the data set used to calculate the \textit{past cast} model performance.
 

### 2017-06-06 letter-editor.tex
Revision 1: Done

@@
 \documentclass{letter}
+\usepackage{graphicx}
+
 \begin{document}
@@
 
-we have substantially revised the manuscript, which we believe is now suitable for publication.
+We have substantially revised the manuscript, which we believe is now suitable for publication.
+
 Reviewer 1 argued that the presented models 'did not result in any significant advance in the forecast of Dst index'.
@@
 Instead, we have decided to make all our codes available.
-Moreover, we feel that well-established standard metrics on which to evaluate the performance of models are still missing in this community.
+Moreover, we feel that well established standard metrics on which to evaluate the performance of models are still missing in this community.
+
 Finally, a major reason to avoid such comparison is that we agree with Referee 2, that the one-hour ahead is of little practical use, because of its strong persistence behavior.
@@
 In any case, for your eyes only we include here a figure showing the comparison of our models with other published models. Our model outperforms all of them.
+
+\includegraphics[width=\textwidth]{figures/Compare_RMSE.png}
 
